Dissecting infant leukemia developmental origins with a hemogenic gastruloid model

Version 1

Abstract

Abstract

Current in vitro models of developmental blood formation lack spatio-temporal accuracy and weakly replicate successive waves of hematopoiesis. Herein, we describe a mouse embryonic stem cell (SC)-derived 3D hemogenic gastruloid (hGx) that captures multi-wave blood formation, progenitor specification from hemogenic endothelium (HE), and generates hematopoietic SC precursors capable of short-term engraftment of immunodeficient mice upon maturation in an adrenal niche. We took advantage of the hGx model to interrogate the origins of infant acute myeloid leukemia (infAML). We focused on MNX1-driven leukemia, representing the commonest genetic abnormality unique to the infant group. Enforced MNX1 expression in hGx promotes the expansion and in vitro transformation of yolk sac-like erythroid-myeloid progenitors (EMP) at the HE-to-hematopoietic transition to faithfully recapitulate patient transcriptional signatures. By combining phenotypic, functional and transcriptional profiling, including at the single-cell level, we establish the hGx as a useful new model for the study of normal and leukemic embryonic hematopoiesis.

Introduction

Blood formation in the embryo is a multi-stage process that develops across multiple cellular niches, reflecting a complex hierarchy of tissue interactions. It involves sequential production of distinct cell types in so-called hematopoietic waves, separated in time and space (Lacaud and Kouskoff, 2017; Costa, Kouskoff and Lacaud, 2012; Medvinsky, Rybtsov and Taoudi, 2011; Dzierzak and Bigas, 2018). The first wave of hematopoiesis produces unipotent red blood cell and macrophage precursors in the yolk sac (YS) from an angioblast precursor which can also form endothelium; it occurs at mouse embryo day (E)7-E7.5 (Fujimoto et al., 2001; McGrath and Palis, 2005; Moore and Metcalf, 1970). The subsequent wave, known as definitive, produces: first, oligopotent erythro-myeloid progenitors (EMPs) in the YS (E8-E8.5); and later myelo-lymphoid progenitors (MLPs – E9.5-E10), multipotent progenitors (MPPs – E10-E11.5), and hematopoietic stem cells (HSCs – E10.5-E11.5), in the aorta-gonad-mesonephros (AGM) region of the embryo proper (Ivanovs et al., 2011; McGrath et al., 2015; Palis et al., 1999; Medvinsky and Dzierzak, 1996; Medvinsky, Rybtsov and Taoudi, 2011). Definitive blood waves are characteristically specified from a specialised hemogenic endothelium (HE) through a process of endothelial-to-hematopoietic transition (EHT) (Zovein et al., 2008; Li, W. et al., 2005). HSC specification from HE is the last event in hematopoietic development and occurs in characteristic cell clusters found on the ventral wall of the dorsal aorta (Ivanovs et al., 2011; Boisset et al., 2010). All blood cell types enter circulation from their original locations and migrate to the fetal liver (FL) (Ciriza et al., 2013; Ghiaur et al., 2008). Most embryonic blood progenitors are transient and sustain blood production exclusively during embryonic development, through differentiation. In contrast, HSCs make little or no contribution to embryonic and fetal blood: instead, they proliferate (i.e. self-renew) to expand their numbers. At least a subset of HSC migrates to the bone marrow (BM) niche at the end of gestation and balances self-renewal and differentiation to maintain blood production throughout the entire post-natal life (Bowie et al., 2006; Kikuchi and Kondo, 2006; Ganuza et al., 2022; Yokomizo et al., 2022).

Blood progenitors and HSCs can be targeted by genetic alterations that confer clonal advantage and/or drive initiation and development of leukemia. Leukemia is the most common malignancy and the most common cause of cancer-related death in the paediatric age group (Steliarova-Foucher et al., 2017; Dong et al., 2020). Several paediatric leukemias originate in utero, and a subset of these are driven by mutations exclusive to the paediatric group (Masetti et al., 2015; Cazzola et al., 2021). Such pediatric-exclusive mutations target blood cell types restricted to development or depend on signals that are specific to embryonic hematopoietic niches, effectively defining distinct biological entities to adult malignancies (Cazzola et al., 2021; Bolouri et al., 2018; Chaudhury et al., 2018). Dissection of these forms of developmental leukemia is confounded by the inability to robustly identify and isolate their embryonic cells of origin, and/or to fully recapitulate the disease in post-natal cells, which leads to failure in generating relevant targeted therapies (Alexander and Mullighan, 2021). One such example is t(7;12)(q36;p13) acute myeloid leukemia (AML) driven by rearrangement and overexpression of the MNX1 gene locus: MNX1-rearranged (MNX1-r) AML is restricted to infants under the age of 2 and carries a high risk of relapse and poor survival (Ragusa et al., 2023). MNX1 is a developmental gene required for specification of motor neurons and pancreatic acinar cells (Thaler et al., 1999; Harrison et al., 1999). MNX1 is not known to participate in embryonic haematopoiesis, however it has been discordantly described to be expressed in some subsets of adult haematopoietic cells (Deguchi and Kehrl, 1991; Nagel et al., 2005; Wildenhain et al., 2012; Taketani et al., 2008). It is ectopically expressed in AML cells carrying one of a variable set of t(7;12) translocations which place the normally silent MNX1 gene under the control of ETV6 regulatory elements (Weichenhan et al., 2023; Bousquets-Muñoz et al., 2024; Ragusa et al., 2023; Ballabio et al., 2009). ETV6 is required for hematopoietic cell specification from HE via VEGF signalling (Ciau-Uitz et al., 2010), potentially positioning MNX1 within a critical embryonic hemogenic regulatory network.

Faithful in vitro recapitulation of developmental blood cell types and their embryonic niches has the potential to dissect the cellular and molecular mechanisms presiding at initiation of developmental leukemia, reveal prognostic biomarkers, and indicate therapeutic vulnerabilities which are otherwise challenging to test in the embryo. Despite widespread use of embryonic stem (ES) cell and induced pluripotent stem (iPS) cell-based models of blood specification, a system which recapitulates the spatial and temporal complexity of embryonic blood formation within the respective time-dependent niches is still lacking (Dijkhuis et al., 2023). In the last few years, gastruloid models have emerged as robust avatars of early development, self-organising processes of symmetry breaking, elongation, multi-axis formation, somitogenesis and early organogenesis, with remarkable similarity to embryo processes in space and time (van den Brink, Susanne C and van Oudenaarden, 2021; Beccari et al., 2018; van den Brink, Susanne C et al., 2020). Gastruloids are grown individually in a multi-well format that facilitates tracking and screening of developmental processes. Relevant to blood formation, a variant gastruloid culture system, which successful organises a heart primordium (Rossi et al., 2021), was also able to recapitulate YS blood formation with erythroid and myeloid cellularity (Rossi et al., 2021; Rossi et al., 2022).

Herein, we describe a new protocol of gastruloid formation, i.e. hemogenic gastruloids (hGx), that captures YS and AGM-like waves of definitive blood progenitor specification from HE. The AGM-like wave includes cells capable of short-term hematopoietic engraftment of the spleen and BM of immunodeficient mice, overall suggesting nearly full EMP-to-MPP/HSC recapitulation of embryonic blood development. We harnessed the embryonic haematopoietic context provided by hGx for modelling MNX1-r infant leukemia by proxy of MNX1 overexpression. MNX1-overexpressing hGx showed expansion of cells at the HE-to-EMP transition, which became susceptible to transformation in vitro and recapitulated MNX1-r AML patient signatures, placing MNX1’s leukemogenic effects at a very early stage of developmental blood formation.

Results

HGxs capture cellularity and topography of developmental blood formation

With the aim of recapitulating embryonic haematopoiesis with spatial and temporal accuracy, we modified existing gastruloid formation protocols to promote the specification of hemogenic mesoderm derivatives and extend developmental time beyond the E8.0-equivalent of early developmental gastruloids (Van den Brink, Susanne C et al., 2014) and the E9.0 timepoint of cardiogenic gastruloids (Rossi et al., 2022). We established a 216h hGx protocol (Fig. 1A) using a Flk1-GFP mouse ES cell line (Jakobsson et al., 2010), which allows the tracking of early hemato-endothelial specification by expression of the Flk1 (i.e. Kdr) locus. Activation of the Flk1 expression, visible by fluorescence of the GFP tag, necessitated induction of the TGF-β signalling pathway by Activin A, in addition to the characteristic gastruloid patterning via WNT activation by CHI99021 supplementation at 48h (Fig. S1A). We sought to promote hemato-endothelial programmes through addition of VEGF and FGF2 at 72h (Sroczynska et al., 2009) (Fig. 1A), and observed extension of a polarised Flk1-GFP-expressing cellular network (Fig. 1B), which included VE-cadherin+ C-Kit+ cells suggestive of EHT (Fig. 1C-D). This process of HE specification and EHT peaked at 120-144h; it was followed by a surge in CD41+ candidate hematopoietic cells at 144h (Fig. 1D). In our experience, observation of early polarised activation of the Flk1 locus at 96h (Fig. 1B) was predictive of CD41+ detection 2 days later, suggesting that TGF-β activation in early patterning is required for later hematopoietic development, putatively by modifying mesoderm specification. We extended the protocol by incorporation of a 24-hour pulse of sonic hedgehog (Shh) between 144-168h to mimic the aortic patterning that precedes cluster formation (Fig. 1A) (Rybtsov et al., 2014). Continued exposure to VEGF after the Shh pulse resulted in specification of CD45+ hematopoietic cells progenitors (Fig. 1D-E), which increased after 168h. The proportion of CD45+ cells was enhanced by the addition of SCF, FLT3L and TPO (Fig. S1B-C), a combination of cytokines normally used for HSC maintenance; conversely, it did not require continued addition of FGF2 after 168h (Fig. S1D), which was omitted. CD45+ cells appeared in a small number of ‘cluster’-like structures adjacent to CD31+ endothelium (Fig. 1E), which were reminiscent of HSC and progenitor cell budding from the aortic HE, and suggested recapitulation of hemogenic tissue architecture. Time-dependent specification of C-Kit+, CD41+ and CD45+ cells in hGxs could be obtained with other mouse ES cell lines, including commonly used E14Tg2a (E14) cells (Fig. S1E-G), with successful specification of comparable frequencies of CD45+ progenitors (Fig. S1H), attesting to the reproducibility of the protocol.

Hemogenic gastruloids (hGx) produced from mES cells promote hemato-endothelial specification with spatio-temporally accurate ontogeny.

(A) Timeline of mES cells assembly and culture into gastruloids over a 216h period with the addition of specified factors for the promotion of hemato-endothelial specification. The 24-hour pre-treatment in 2i+LIF was omitted when mES cultures showed no signs of differentiation. (B) Imaging of hGx over time from 72h to 216h at 10x magnification, showing the assembly and growth of the 3D structures and the polarization of the Flk-1-GFP marker from 96h; scale bar: 100mm. (C) Flow cytometry timecourse analysis of hGx for the presence of CD144 (VE-cadherin), C-Kit, CD41, and CD45 markers; representative plots. (D) Quantification of analysis in (D). Violin plots with median and interquartile range of up to 26 replicates; Kruskal-Wallis CD144 (p=0.0755), C-Kit (p<0.0001), CD41 (p<0.0342), CD45 (p<0.0001). Shown are significant Dunn’s multiple comparisons tests; adjusted q-value. (E) Immunostaining of whole individual gastruloids at 216h showing the localized expression of CD31 (yellow), C-Kit (turquoise), and CD45 (pink) markers; scale bar: 100μm.

Overall, our extended hGx model could capture successive specification of HE, CD41+ and CD45+ cells, with temporal and spatial congruence relative to the embryo, thus showing initial promise as a tool for the exploration of normal and perturbed developmental hematopoiesis.

To better characterize the extent and progression of developmental hematopoiesis in hGx, we performed single-cell RNA-sequencing (scRNA-seq) time-course analysis of gastruloid cell specification. We sorted cells from 2 independent hGx cultures at 120, 144, 168, 192 and 216h, and profiled a total of 846 cells using the Smart-Seq2 protocol (Picelli et al., 2014) (Fig. S2A). We analysed unfractionated live single hGx cells at 120-192h and enriched the dataset with sorted CD41+ cells at their peak time of 144h, and sorted CD45+ cells at 192 and 216h. At timepoints of significant enrichment in CD41+ (144h) and CD45+ cells (192h) (Fig. 1D), we also sorted C-Kit+ cells. Library preparation and sequencing generated an average of 120000 reads/cell, which were mapped to an average of 4000 genes/cell, with minimal signs of cell stress / death as measured by mitochondrial DNA fraction (Fig. S2B). Read and gene counts were similar between biological replicates, cell types and at different time points (Fig. S2C). The exception was 120h unfractionated hGx cells, which despite similar sequencing depth (average read counts), were mapped to twice the number of genes (Fig. S2C), possibly reflecting multi-gene program priming at the onset of hemogenic specification. We selected highly varying genes (HVG) before principal component analysis (PCA) dimensionality reduction and retained the most relevant dimensions. In the PCA reduced space, we constructed a KNN graph and used uniform manifold approximation and projection (UMAP) to visualize the data on 2 dimensions, looking for cell communities using Leiden clustering (Fig. 2A). We identified 12 cell clusters of which 2 (clusters 0 and 5) almost exactly mapped to C-Kit+ cells (also Sca-1+), 2 (clusters 1 and 8) contained CD45+ sorted cells, and cluster 4 uniquely captured the CD41+ cells observed at 144h (Fig. 2A). Although some unfractionated cells overlapped with the hemogenic cell clusters (Fig. 2A - center), most occupied different transcriptional spaces, reflecting the frequency of hematopoietic cells and suggesting the presence of potential hemogenic niches.

Time-resolved analysis of scRNA-seq of hGxs captures successive waves of hematopoietic specification.

(A) UMAP projection of all sequenced cells colored by annotated clusters (left), by sorting markers C-Kit/ScaI, CD41, and CD45, or unfractionated cells (‘whole’ corresponding to ‘all’ in panel A) (center), and by hGx culture time (right). (B) Heatmap representation of the expression of endothelial and hematopoietic marker genes in individual hGx C-Kit+ cells sorted at 144 and 192h. Cells and genes ordered by unsupervised hierarchical clustering; cell cluster dendrogram in Fig. S2E. (C) Heatmap representation of the expression of endothelial and hematopoietic marker genes in individual hGx CD41+ and CD45+ cells sorted at 144 and 192/216h, respectively. Cells and genes ordered by unsupervised hierarchical clustering; cell cluster dendrogram in Fig. S2F. (D) Summary representation of the proportion of hGx cells expressing endothelial (endo), erythroid/myeloid (Ery+My), myeloid/lymphoid (My+Ly) and HSPC gene signatures. Endo: Kdr; Ery+My: Epor±Gata1±Klf1±Hbb±Hba±Hbg + Spi1±Mpo±Anpep±Csf1/2/3r); My+Ly: Spi1±Csf1/2a/3r + Ikzf1±Ighm/d±Igk); HSPC: Ptprc + Myb). (E-G) Cell type-associations of selected surface markers and transcription factors affiliated with the HE-to-hematopoietic transition (CD44, Procr, Etv6, Gata2, Notch1, Tal1) and with hematopoietic (Gfi1b, Mllt3, Runx1) and definitive hematopoietic (Ikzf1, Hlf, Hoxa9, Myb) gene expression programs with (E) C-Kit+ (endo-haem) and CD41+/CD45+ (haem), (F) endo-haem cells obtained at 144h or 192h, and (G) CD41+ or CD45+ cells. Gene expression analysed as present (non-zero expression) or absent (zero-expression). Odds ratio with significance (-log10p<1) estimated by Fisher’s test.

To explore tissue and lineage affiliation of the different clusters, we performed differential gene expression against all other cells using Wilcoxon ranking test, and established cluster classifier gene lists (Supplemental File S1). Top classifier genes for clusters 0 and 5 had endothelial affiliation (Supplemental File S1), and the cellular composition of the clusters broadly reflected their time signature (Fig. 2A): cluster 5 comprised 144h cells and cluster 0 included most C-Kit+ cells at 192h. Cluster 5 occupied the vicinity of 144h-CD41+ cells (Fig. 2A), which carried an erythroid-biased signature (Supplemental File S1). On the other hand, 192/216h-CD45+ cells in cluster 8 expressed myeloid and lymphoid genes (Supplemental File S1). The remaining clusters largely corresponded to time-restricted unfractionated cells (Fig. 2A - right), with clusters of cells present at later time-points matching mesenchymal stromal cell and autonomic neuron signatures (Fig. S2D), potentially configuring incipient formation of niches relevant to production of HSC (Kapeni et al., 2022; Fitch et al., 2012).

To further characterise the identity of hGx cells, we interrogated lineage priming in individual hemato-endothelial cells (Fig. 2B-C), to probe the presence of putative multilineage YS (EMP) and AGM (MLP/MPP) progenitors, and HE cells. The heatmaps (Fig. 2B-C) show expression of lineage markers and regulators, with unsupervised clustering of cells (Fig. S2E-F) and genes. The gene list used was identical in both analyses, with only expressed genes shown. Concordance between sorting antigens and expression of the respective genes (Kit, Itga2b [CD41] and Ptprc [CD45]) was partial potentially reflecting low levels of transcript expression in the hGx system. Summary of endothelial, co-expressing erythro-myeloid (EM) and co-expressing myelo-lymphoid (ML) programs (Fig. 2D) clearly associates the first with C-Kit+ and the latter with CD45+ cells; CD41+ cells capture a heterogeneous population, which includes a proportion of cells co-expressing erythroid and myeloid genes that constitute candidate EMPs.

Clustering of cells sorted as C-Kit+ did not separate the 144h and 192h time-points (Figs. 2B and S2E), indicating that the relative separation on the global UMAP was not based on endothelial identity genes. We observed widespread expression of arterial genes (e.g. Flt1, Efnb2) with relative reciprocity with venous endothelial genes (e.g. Flt4) supporting specification of both types of endothelia (Fig. 2B). There was limited co-expression of hematopoietic genes, but genes associated with endothelial-to-hematopoietic transition (EHT), e.g. Gata2, Notch1, and Procr, were significantly more frequent within the C-Kit+ population (Fig. 2E). Specific associations of Cd44 and Procr with the 192h time-point (Fig. 2F) putatively represent aorta-like EHT.

Clustering of CD41+ and CD45+ cells (Fig. 2C and S2F) more clearly separates unique hematopoietic signatures. A subset of cells, predominantly CD41+, expresses high levels of globin genes. While embryonic β-globin chains are highly expressed, there is a gain in fetal Hbb-y and adult Hbb-bt and Hbb-bs (C57Bl/6 strain-specific β major and β minor, respectively) in CD45+ cells at 216h, indicative of AGM-like progenitor identity (Fig. 2C). 192 and 216h cells with high expression of Ptprc have an MLP profile (Fig. 2C-D), co-expressing myeloid and lymphoid genes including immunoglobulin chains. There is progression of the lymphoid signature between 192 and 216h (Fig. 2D and S2G). Other CD45+ cells have low-frequency expression of HSC-associated transcription factors, including Myb, Hlf and Hoxa9 (Fig. 2G and S2I-J), suggesting incipient specification of AGM-like MPP and pre-HSC, albeit with incomplete programming.

Altogether, single-cell RNA-seq data suggest that hGx specifies YS-like and AGM-like progenitors in a time-dependent manner, with concomitant emergence of hematopoietic-supporting niche cells. The data suggested putative incipient specification of rare candidate MPP or pre-HSC, which we investigated further in comparison with mouse AGM datasets and at a functional level.

We made use of a dataset recently published by one of us (Thambyrajah et al., 2024) that interrogates the mouse AGM at the point of HSC emergence from the dorsal aorta HE. In this study, a combination of surface markers and Gfi1 locus reporting of EHT, tested functionally through transplantation, identified HE and HSC-enriched clusters (Fig. S3A - clusters 1 and 2, respectively, with cluster 4 a putatively heterogenous cluster at the transition between the 2 states). We projected cell transcription profiles from hGx to AGM using a version of scmap algorithm (Kiselev, Yiu and Hemberg, 2018) with 1-cell and 3-cell neighbourhoods for lower and higher levels of confidence on similarity projections (respectively gray and black circles on Fig. S3A - left). The numbers of cluster-specific hGx cells projected onto AGM clusters is quantified in Fig. S3B. Top differentially expressed genes (DEG) between AGM HE and HSC-projected hGx cells are represented in Fig. S3C and Supplemental File S2.

With respect to HE specification, we observed that a subset of hGx C-Kit+ cells present in endothelial clusters 5 and 0 (Fig. 2A) projected onto the AGM HE-identity cluster 1 (Fig. S3A), with more frequent similarity amongst hGx 192h cluster 0 (41%) than 144h cluster 5 (21%). Overall, this supports the notion that specification of endothelial and HE cells in hGx follows with time-dependent developmental progression into putative AGM-like HE. Lineage identity genes supported the nature of HE-projected hGx cells (Fig. S3D), which expressed Gata2 but not Runx1, Myb, or Gfi1b, indicating that EHT occurs downstream of these cells. We then asked if any of the hGx cells in hemopoietic clusters bore resemblance to cells from the HSC-enriched AGM cluster 2. We observed that most (85%) hGx CD45+ cells in cluster 8 (Fig. 2A) projected onto the HSC-enriched cluster 2, which also showed similarity with 76% of CD41+-enriched hGx cluster 4 cells. The relative positioning of hGx cells cluster 4 and cluster 8 progressively further away from the AGM HE cluster was in line with the time-dependent specification in the hGx. However, the lineage marker programs (Fig. S3D) expressed by the hGx cells, although definitive in nature (e.g. there is expression of immunoglobulin chains and adult β-globins), lacked detection of HSC genes like Hlf, Gfi1, or Hoxa9. On the other hand, Mllt3 was present (Calvanese et al., 2019), even in the absence of an erythroid signature (Pina et al., 2008), suggesting a developing pre-HSC program compatible with the co-expression of multi-lineage markers (Fig. S3D).

Late-stage hGx cells can engraft hematopoietic tissues upon maturation

We inspected the cellularity of hGx at the 216h time-point more closely through a combination of multi-parameter flow cytometry (Fig. 3A-B), colony-forming progenitor assays (CFC) (Fig. 3C-D) and in vivo engraftment of haematopoietic tissues (Fig. 3E-G). In order to facilitate the analysis of engrafted animals, we engineered constitutive expression of BFP from the Rosa26 locus in the Flk1-GFP mES cell line (Fig. S4A), which we also used in some of the in vitro analyses. Phenotypic characterisation of 216h-hGx (4 replicates, 1 with Rosa26-BFP::Flk1-GFP line) (Fig. 3A, S4B) showed the presence of pre-HSC-like CD45+CD41loc-Kit+VE-Cadherin+ cells (1.8% of all CD45+CD41lo cells) (Fig. 3B), corresponding to an estimated 8 cells per hGx. Analysis of the hGx progenitor content using multipotential CFC assays (Fig. 3C-D) revealed the presence of >50 CFC/hGx, of which an average of 7 have mixed-lineage granulocytic-monocytic-erythroid (GEM) potential, numerically matching the estimated number of pre-HSC. Other progenitors were predominantly granulo-monocytic (GM) and erythroid (E) progenitors (Fig. 3C-D). Differentiated cells of G, M and E lineages (Fig. S4C-D) were also detectable in 216h hGx, putatively reflecting the progeny of EMP-like cells specified earlier at the 144h timepoint.

HGxs specify cells with pre-HSC characteristics.

(A) Flow cytometry analysis of 216h Rosa26-BFP::Flk1-GFP hGx cells for CD41, CD45, c-Kit and CD144 (VE-Cadherin). (B) Quantification of pre-HSC-like c-Kit+CD144+ cells within the CD45+CD41lo fraction of 216h-hGx; mean of 4 independent replicate experiments. (C) Colony-forming cell (CFC) assay of 216h-hGx in multipotential methylcellulose-based medium. GEM: granulocyte-erythroid-monocyte; GM: granulocyte-monocyte; M: monocyte; E: erythroid. Please note that the medium does not include TPO and does not assess the presence of megakaryocytic progenitors. CFC frequency in 3hGx, n=3, mean±SD. (D) Representative photographs of colonies quantified in C; scale bar=2mm. (E) Schematic representation of the experimental workflow for execution and analysis of hGx implantation in the adrenal gland of immunodeficient mice. (E) Histology of implanted and control adrenal glands; paraffin sections stained with hematoxylin & eosin (H&E). Inserts in implanted animal highlight neural (orange) and hematopoietic and renal tubular epithelium (green); inserts in control animal highlight normal cortical (blue) and medullar (red) adrenal architecture, absent in implanted adrenal glands. (F) PCR detection of hGx genomic (g)DNA in the bone marrow (BM) and spleen (Spl) of immunodeficient mice 1 month after unilateral adrenal implantation of 3 gastruloids/gland. Analysis of 5 replicates of 100ng gDNA /recipient tissue; control animal was injected unilaterally with PBS in an adrenal gland in parallel with experimental implantation. Reaction positive control used 100ng of gDNA from Rosa26-BFP::Flkj1-GFP mES cells (mES) used to generate hGxs. NTC: no template control. See Fig. S4F-H for additional details.

As an ultimate test of the presence of pre-HSC, we tested whether late-stage 216h hGx contained cells capable of hematopoietic engraftment of immunodeficient mice. In early experiments, we transplanted dissociated hGx cells into irradiated immunocompetent C57Bl/6 recipients, sub-lethally irradiated immunodeficient NSG mice, or non-conditioned NSGW41 hosts, but failed to detect signs of engraftment of peripheral blood, spleen (Spl) or bone marrow (BM) at 12 days (including absence of spleen colony formation), at 4-6 weeks (short-term engraftment), or beyond 8-10 weeks (long-term engraftment) by either flow cytometry detection of CD45.2 or by genomic DNA (gDNA) PCR analysis of the Flk1-GFP locus. However, given the absence of clear HSC-like transcriptional programs in 216h CD45+ cells, as analysed by scRNA-seq (Fig. 2 and S2), and in light of the late specification of mesenchymal or sympathetic neuronal cells (Fig. 2A and S2D) putatively supportive of HSC formation, we hypothesized that hGx pre-HSC required additional maturation in a relevant supportive niche. We thus tested implantation of undissociated hGx in the adrenal gland of Nude mice (Fig. 3E), a topography commonly used to support tumor development, and recently shown to support adult haematopoietic development (Schyrr et al., 2023). Given the coincidence of CD45 isoform between Flk1-GFP mouse ES cells and Nude mice, we made use of the Rosa26-BFP::Flk1-GFP mES cell line (Fig. S3A). We implanted 3 hGx each unilaterally in the adrenal gland of unconditioned recipient mice. Control mice were injected with an equivalent volume of PBS. Animals were sacrificed at 24 hours, 12 days, and 4 weeks after implantation, with 5 experimental animals and 1 control collected at each timepoint. We checked Spl and BM BFP engraftment at the different timepoints and analyzed the morphology of adrenal glands receiving implants (Fig. 3F). We did not observe BFP engraftment, spleen enlargement or colony development, or macroscopic changes in the adrenal gland at the earlier timepoints. However, 3 of 5 adrenal glands were enlarged, and their histology suggested in situ development of the hGx (Fig. 3F), particularly mesodermal tissues – somite derivatives (Fig. 3F - left, red box), and lateral plate and intermediate mesoderm derivatives, respectively blood cells and renal tubules (Fig. 3F - left, green box). The control, PBS-injected glands showed normal tissue architecture (Fig. 3F - right). It should be noted that none of the specimens contained glomeruli in addition to the renal tubules, pointing to an origin from the hGx, and not from the adjacent kidney tissue. BFP engraftment of the Spl and BM by flow cytometry was very low level albeit consistently above control (Fig. S4E), and we opted for analysing engraftment by PCR analysis of gDNA using primers directed against the BFP insert (Fig. S4A). We measured the sensitivity of the amplification through serial dilution of Flk1-GFP Rosa26-BFP mES cell gDNA into that of human K562 cells, and determined detection of 10pg/100ng gDNA, equivalent to 2-3 cells (Fig. S4F). PCR analysis of spleen and BM gDNA detected BFP amplicon in both tissues in 3 of the 5 samples (#11, 12 and 15) (Fig. 3G). We sampled the gDNA of each recipient 4-5 times and detected the presence of the amplicon in only a fraction of the replicate samples of the positive animals, reflecting the low-level detection by flow cytometry. Replicate sampling of the control was consistently negative (Fig. 3G). Sorting of erythroid Ter119+ and lympho-myeloid CD45+ cells (Fig. S4G) showed PCR engraftment in both lineage fractions (Fig. S4H).

Altogether, the data suggest that 216h hGx generate AGM-like pre-HSC capable of at least short-term multilineage engraftment upon maturation. The low-level production of engrafting cells recapitulates their rarity in vivo, in agreement with the embryo-like qualities of the gastruloid system, which accommodates modular qualitative bias, but not quantitative changes in tissue specification.

The ability of hGx to reconstitute YS and AGM-like HE and hemopoietic progenitors with temporal coherence lends itself as an attractive model to understand developmental leukemia. In particular, forms of AML unique to the first 2 years of life – infant (inf)AML – have been shown to transform fetal liver (FL) but not adult haematopoietic cells (Waraky et al., 2024; Mercher et al., 2009; Chen, W. et al., 2011; Ragusa et al., 2023), compatible with targeting of a transient embryonic cell. In the case of t(7;12) AML (herein MNX1-r) (Ragusa et al., 2023), which results in a transformation-driving ectopic expression of MNX1 (Waraky et al., 2024), analysis of patient transcriptional signatures (Ragusa et al., 2022) for over-representation of cell atlas-defined affiliations, indicates enrichment in cardiomyogenic, endothelial, HSC/progenitor and mast cells (Fig. 4A). These cell enrichments are compatible with an early hemogenic cell derivative, capturable in the hGx model. Other cell type-enrichments reflect MNX1 functions in pancreatic and neuronal development (Ragusa et al., 2022) (Fig. S5A).

HGxs with MNX1 overexpression have increased cellularity and enhanced hemogenic endothelial potential.

(A) Cell type enrichment analysis in MNX1-r AML transcriptomes, compared to normal paediatric bone marrow (BM) or other paediatric AML from the TARGET database. GSEA used representative cell type gene sets from the 2021 DB database; bubble plot shows NES scores (color gradient) and statistical significance (– log10(FDR), bubble size). (B) Schematic representation of generation of MNX1-overexpressing and empty vector (EV) mES cells by lentiviral transduction, with subsequent assembly into hGxs (haemGX). (C) Imaging of hGxs with MNX1 overexpression and EV controls at 10x magnification, showing appropriate assembly and polarization of the Flk1-GFP marker. scale bar: 300mm. (D) Size of hGx at each timepoint, determined by the distance between the furthest extreme points in µm. Mean ± SD of 3 replicate experiments; 2-tailed t-test, p< 0.05 (*), 0.001 (**), 0.0001 (***), and 0.00001 (****). (E-G) Flow cytometry timecourse analysis of (E) C-Kit+, (F) CD41+, and (G) CD45+ cell abundance in MNX1 and EV hGxs (120-216h). Mean ± standard deviation of 3-7 independent experiments; 2-way ANOVA and Sidak’s multiple comparison test significant at p<0.05 for C-Kit+ cells only (construct contribution to variance p=0.0191; 144h comparison *p=0.0190). (H) Volcano plots of differentially expressed genes (DEGs) by bulk RNA-seq between MNX1 and EV hGxs at 144h and 216h. Intersection of statistically significant DEGs between 144 h and 216 h (shown in Venn diagram on top right) identified unique DEGs for each condition (labelled in red for 144 h and blue for 216 h). (J) Gene Ontology (GO) term enrichment for differentially upregulated genes between MNX1 and EV hGx at 144h (red) and 216h (blue), computed in EnrichR using the GO Biological Process repository. (I) Transcription factor (TF) binding site enrichment on differentially upregulated genes between MNX1 and EV hGx at 144h (red) and 216 hr (blue) using the ENCODE and ChEA Consensus TFs from ChIP database on EnrichR.

We overexpressed MNX1 (MNX1-OE) in Flk1-GFP mouse ES cells by lentiviral transduction (Fig. 4B) and analyzed progression of hGx generation (Fig. 4C). We used human MNX1 cDNA to distinguish from the endogenous gene, but the degree of homology is nevertheless high (84%), supporting functional equivalence. We confirmed that MNX1 overexpression in hGx is maintained at endpoint (Fig S5B). MNX1-OE hGx activated polarized Flk1-GFP expression and elongated with similar kinetics to empty-vector (EV) control (Fig. 4C), but consistently produced larger hGx (Fig. 4D) denoting increased cellularity (Fig. S5C). From 192h onwards, MNX1-OE hGx had a higher frequency of spontaneously contractile structures (Fig. S5D), compatible with the cardiogenic cell association of patients’ transcriptomes (Fig. 4A). We interrogated the time-dependent hemogenic cell composition of hGx by flow cytometry, quantifying C-Kit, CD41 and CD45 (Fig. 4E-G and S5E-F). We observed a significant expansion of the C-Kit+ compartment specifically at 144h (Fig. 4E and S5E). All cell markers were relatively unchanged at later timepoints, although absolute cell numbers were higher in MNX1-OE hGx (Fig. S5C).

To better understand the consequences of MNX1-OE on hemogenic development, we performed RNA-sequencing (RNA-seq) of hGx at 144h and 216h, matching YS and AGM-like hemopoietic waves, respectively (Fig. 4H). We confirmed expression of the human MNX1 transgene in the sequenced reads (Fig. S6A); in contrast, we could not detect endogenous mouse Mnx1 in any of the samples (Fig. S6A), confirming that it does not normally play a role in hemogenic development. DEGs between MNX1-OE and EV (Supplemental File S3) configured distinct MNX1-driven programs at 144h and 216h, with >80% genes timepoint-specific (Fig. 4H). Both programs include up-regulation of hematopoietic-associated genes, capturing EHT and progenitor regulators (e.g. Runx1, and Epor and Zfpm1, respectively) at the earlier timepoint, and more differentiated effectors, e.g. Hba-a1, Hbb-bh1 and Il3ra at 216h, compatible with selective targeting and/or expansion of newly specified progenitors at the YS-like stage (Fig. 4H). An overview of transcriptional regulatory programs at both timepoints denotes a clearer hemogenic enrichment at 144h, with over-representation of GATA2, GATA1 and RUNX1 binding targets (Fig. 4I), putatively capturing an expansion of the HE-to-EMP transition, also represented in enriched GO categories related to VEGF signalling and epithelial-to-mesenchymal transition (EMT) (Fig. 4J). Compatible with the increase in contractile foci in MNX1-OE late hGx, there was an enrichment in cardiogenic gene ontologies (GO) at the 216h timepoint (Fig. 4J). No hemogenic transcription factors (TF)-ChIP or GO categories were enriched in MNX1-OE hGx at 216h (Fig. 4J-I).

We attempted to deconvolute the cellular heterogeneity underlying bulk RNA-seq DEGs by interrogating the cluster signatures obtained from the single-cell analysis of hGx development (Fig. S6B-C; refer to Fig. 2A). In support of enhanced YS-stage hemato-endothelial specification, MNX1-OE transcriptomes showed enrichment of hGx HE cluster 5 and 0 signatures, and of the EMP-like cluster 4 (Fig. S6B). MNX1-OE DEGs at both timepoints also enriched for the relatively less differentiated cluster 10, which predominantly captures unfractionated hGx cells at 120h (Fig. S6B-C; refer to Fig. 2A). Significantly, clusters 5 and 0 HE signatures were enriched in MNX1-r infAML patient signatures (Ragusa et al., 2022) (Fig. S5B-C), positioning MNX1-OE at early stages of hemogenic specification. MNX1-OE DEGs at 216h show enrichment in late-stage clusters 7 and 8 (Fig. S6B-C), denoting enhancement of hGx development, including MLP/MPP in cluster 8, but diverging from MNX1-r infAML signatures. Projection of infAML subtype-specific gene signatures onto hGx development separates the lineage affiliations of MNX1-r from, for example, the most frequent infAML driver – KMT2A (MLL) rearrangements (Fig. S6B-C), which carries myelo-monocytic and/or mixed-lineage affiliation, highlighting the utility of the hGx system for interrogating infAML biology.

We further interrogated the MNX1-r-like leukemia-initiating potential of MNX1-OE hGx through serial replating of CFC assays, a classical in vitro assay of leukemia transformation. We compared 144h and 216h hGx to match functional and transcriptome data, and plated unfractionated hGx, given the transient nature of C-Kit+ enrichment. Both 144h (Fig. 5A-B) and 216h (Fig. 5C-D) MNX1-OE hGx replated for at least 5 plates, with significant differences to control (EV) (Fig. 5B,D). EV-hGx replating could not be sustained beyond plate 3 in most experiments started from 216h (Fig. 5C); 144h EV-hGx exhibited some low-level replating until plate 5 (Fig. 5A), but the colonies were formed by few scattered cells clearly distinct from the well-demarcated colonies obtained from MNX1-OE hGx. We analyzed early-stage (plate 1) and late-stage (plate 5) colonies by flow cytometry to identify the nature of the replating cells. Cells selected through replating were C-Kit+ (Fig. 5E), matching the cell phenotype transiently enriched at 144h. Early replating of MNX1-OE hGx also enriched for erythroid-affiliated Ter119+ cells particularly from the 144h time-point (Fig. 5F), reflecting the differential erythroid potential of EMP and MLP/MPP at 144h and 216h, respectively. This enrichment is not sustained upon replating, compatible with the undifferentiated nature of MNX1-r infAML, which is also reflected by the low levels of CD11b, a marker affiliated with the myelo-monocytic lineages (Fig. 5E). Notably, the cells obtained are not CD45+ (Fig. 5F), suggesting perpetuation of an early hemogenic state.

MNX1 selects C-Kit+ clonogenic cells and selectively transforms end-stage hGxs.

(A) Representative photographs of serial replating of colony-forming cell assays initiated from EV or MNX1 cells obtained at 144h of the hGx protocol. (B) Quantification of colony-replating efficiency of EV and MNX1 144h-hGx cells. Mean+SD of n>3 replicates; Kruskal-Wallis with Dunn’s multiple comparison testing at significant q<0.05. (C) Representative photographs of serial replating of colony-forming cell assays initiated from EV or MNX1 cells obtained at 216h of the hGx protocol. (D) Quantification of colony-replating efficiency of EV and MNX1 216h-hGx cells. Mean+SD of n=5-8 replicates; Kruskal-Wallis with Dunn’s multiple comparison testing at significant q<0.05. (E) Flow cytometry analysis of hemogenic progenitor C-Kit+ cells and myeloid-affiliated CD11b+ cells at first round of plating of 144h and 216h-hGxs initiated from EV and MNX1 cells; representative plots. (F) Representative flow cytometry plots of serially re-plating MNX1 cells from 144h and 216h-gastruloids (plate 5 cells).

We performed RNA-seq analysis of MNX1-OE hGx CFC cells obtained through re-plating and integrated the data with the 144h and 216h hGx time-points to identify genes enriched through the process of transformation. We observed relative enrichment in HsMNX1 reads in MNX1-OE CFC samples, supporting selective expansion of transduced cells (Fig. 6A). Hierarchical clustering of all DEGs in pairwise comparisons (Fig. 6B), identified a subgroup of genes up-regulated upon CFC replating which was also up-regulated at MNX1-OE hGx at 144h (K14), thus matching the putative time-dependent enrichment of the MNX1-OE propagating population. To understand the relevance of these genes in MNX1-r leukemia, we performed Gene Set Enrichment Analysis (GSEA) of MNX1-r patients in comparison to other AML in the same infant (0-2) age group. K14-genes were enriched in MNX1-r all comparisons (Fig. 6C), with considerable overlap between leading-edge (i.e. significantly-enriched) genes (Fig. S6D). Interrogation of cell-type-specific signatures within the K14 MNX1-r leading edge genes (Fig. 6D) against the single-cell PanglaoDB atlas captured similarities to hemogenic and non-hemogenic cell types, including interneuron, and other neural, signatures, which reflect MNX1 biological functions (Harrison et al., 1999; Thaler et al., 1999; Ragusa et al., 2022). The most highly significant hemogenic signatures capture early developmental components, such as ‘endothelial cells’, which were specifically up-regulated within MNX1-OE hGx (Fig. S6B). ‘Erythroid cell precursors’ also reflect MNX1-OE hGx effects (Fig. 4E-J). We were intrigued by the ‘mast cells’ affiliation of MNX1-r AML signatures within the hGx K14 cluster, which was more specifically aligned with MNX1-OE hGx isolated at 144h (Fig. S6E). Given the characteristic morphology of these cells, we analysed Giemsa-Wright-stained cytospins of CFC replating assays (Fig. 6E and S6F). Consecutive plates of 144h MNX1-OE-initiated in vitro transformation showed a shift from blast-like at plate 3 to precursor cell morphology with some differentiated mast cells at plate 5 (Fig. S6F). However, the extent of differentiation was greater, and apparent earlier, in 216h-MNX1-OE initiated colonies (Fig. 6E). This suggests that the MNX1-OE-responsive hemogenic C-Kit+ progenitor may experience maturation between 144h and 216h, and is more likely targeted for transformation at the YS EMP stage. The data support the utility of the hGx in dissecting ontogeny of embryonic origin in hematological cancers, allowing us to specifically narrow down the initiating cellular targets of MNX1-r AML.

HGxs with MNX1 overexpression recapitulate MNX1-Acute Myeloid Leukemia patient signatures.

(A) Quantification of human MNX1 transcripts in FPKM units from RNA-seq of GX-EV, GX-MNX1, and CFC at 144 hr and 216 hr. Bars represent mean of replicates. (B) Heatmap comparing the expression of all differentially expressed genes (DEGs) between all conditions (GX-MNX1 144 hr vs GX-EV 144 hr; GX-MNX1 216 hr vs GX-EV 216 hr; CFC MNX1 144 hr vs GX MNX1 144 hr; CFC MNX1 216 hr vs GX MNX1 216 hr) as Z score. Hierarchical clustering by Ward D method on Euclidean distances identifies 14 clusters (k). (C) GSEA plots for gene set extracted from k14 from (B) against MNX1-r patients RNA-seq counts vs MLL, core-binding factors (CBF), or other pediatric AML. (D) Cell type analysis of the intersect of leading edge genes LEGs (n=476) from Fig. S5D in MNX1-r patients (k14 LEGs) (B) using the Panglao DB 2021 database. (E) Representative Giemsa-Wright stained dissociated CFC replating MNX1 hGx cells from 144h and 216h. Solid black arrows, mast cell precursors; dashed black arrows, mast cells.

Discussion

In this study, we developed and characterized a hGx model from mouse ES cells, which successfully recapitulates YS-like and AGM-like waves of definitive blood progenitor specification, and can engraft adult hematopoietic tissues upon in vivo maturation. We took advantage of the hGx’s nearly full recapitulation of embryonic hemogenic specification, including critical early stages of HE production and EHT to EMP and MLP/MPPs, to interrogate the biology of MNX1-r AML, a poorly characterized form of leukemia exclusive to the infant age group (Ragusa et al., 2023). We showed that MNX1-OE transiently expanded a C-Kit+ cell with hemato-endothelial characteristics at the YS-like stage, and could selectively sustain the phenotype through serial replating of CFC assays. Replating CFCs could be initiated beyond the YS-like stage, albeit at the relative expense of an early differentiation state and of the similarity to patient transcriptome, putatively defining a window of susceptibility to MNX1-driven transformation. The hGx MNX1-OE propagating cell bears transcriptional resemblance to patient MNX1-r AML, overall positioning the hGx as an attractive tractable model in which to study developmental leukemias.

MNX1-r has been difficult to characterize due to a combination of rarity, diagnostic difficulty (the underlying t(7;12) can be missed if not specifically looked for) (Tosi et al., 2015), and the lack of a fusion protein that can be ectopically expressed and tracked through development, as is the case for other translocations unique to infAML (Alexander and Mullighan, 2021). However, a recent paper using MNX1-OE has shown that MNX1 can initiate a transplantable leukemia in immunodeficient mice by targeting FL, but not BM, cells (Waraky et al., 2024). This suggests that ectopic expression of MNX1 transforms an embryonic cell, rather than depends on an embryonic niche. It also suggests that MNX1-OE can cause leukemia as a single genetic hit, although it cannot be excluded that other mutations are required for, and may indeed have contributed to, full transformation in vivo, as observed in a recent report confirming the in utero origin of the leukemia (Bousquets-Muñoz et al., 2024). However, the mouse model does not address the nature of the embryonic cell targeted by MNX1-OE. Like the MNX1-OE hGx cells selected through serial replating, MNX1-OE mouse leukemia are also C-Kit+ (Waraky et al., 2024). Interestingly, clusters of genes up-regulated in the mouse MNX1-OE leukemia (Fig. S7A – K1, 2, 5, 6, and 9) do not robustly separate MNX1-r from other infAML by GSEA analysis (Fig. S7B), and have diverged more from the patient data in their cell-type signature analysis than the original MNX1-OE cells in the FL (Fig. S7C), putatively selecting a more generic program of leukemia transformation. We propose that the divergence from MNX1-r infAML signatures increases from MNX1-OE hGx 144h (YS-like), to 216h (AGM-like) to FL, positioning the origin of the AML at the YS stage, likely at the HE-to-EMP transition. This cell may persist in the hGx culture system to the 216h AGM-like time-point, and migrate to the FL in the embryo, eventually extinguishing through differentiation, as suggested by the MNX1-OE hGx CFC assays. Persistence of the original MNX1-OE target cell will determine the sensitivity to MNX1-OE, and can be addressed in the hGx system through lineage tracing. Whether additional mutations, namely a reported association between MNX1-r and tri(19) (Espersen et al., 2018; Ragusa et al., 2023), contribute to prolonging the undifferentiated state of the candidate YS target cell with increased probability of transformation, or otherwise expand the targeted cell in a more classical second-hit fashion, can potentially also be screened in the hGx system, in combination with transplantation experiments. Finally, although our study has positioned the MNX1-OE target cell within the YS-EMP stage, the exact nature of the cell requires better definition. The mast cell affiliation, transcriptional and by CFC morphology, is more likely to reflect the period of susceptibility to MNX1 and the multilineage potential of the target cell, than the lineage commitment of the latter. Indeed, blast cells in the leukemia mouse model are morphologically undifferentiated (Waraky et al., 2024) and a mast cell morphology has not been described in patients (Ragusa et al., 2023; Espersen et al., 2018). Although adult HSC have mast cell potential in vitro, this is not employed in vivo under physiological conditions (Chia et al., 2023). Tissue resident mast cells are formed in the embryo at 2 periods: the YS EMP stage, which we propose to have targeted with MNX1-OE in the hGx model; and a transient AGM FL-HSC population which does not persist in post-natal life (Chia et al., 2023; Yoshimoto et al., 2022), and that may be incipiently specified at the end of the current hGx protocol. Prospective isolation of these cell types from embryos and over-expression of MNX1 followed by transformation in vitro and in vivo, will precisely define the embryonic time-window of susceptibility to MNX1-r and the nature the leukemia-initiating cell. Harnessing the high-throughput nature of the hGx culture system will help identifying therapeutic vulnerabilities that can improve the prognosis of this often-fatal disease.

The hGx model furthermore constitutes an advanced platform to probe physiological development of the hematopoietic system. Currently, in vitro platforms do not recapitulate in vivo blood formation sufficiently faithfully (Dijkhuis et al., 2023; Fidanza and Forrester, 2021). Establishment of a tractable in vitro model of hematopoietic development which faithfully recapitulates time-dependent emergence of hematopoietic progenitors in their physiological niche would propel translational applications by enabling high-throughput studies of HSC emergence. Most ESC-based models of blood specification do not achieve concomitant recapitulation of the surrounding microenvironment. Protocols typically remove ESC-derived hemogenic cells from an early embryonic-like niche to direct differentiation through genetic manipulation (Sugimura et al., 2017) and/or empirically-determined growth factor addition (Lis et al., 2017; Vo and Daley, 2015; Fowler et al., 2024; Nafria et al., 2020; Sroczynska et al., 2009; Piau et al., 2023), with minimal probing of intermediate steps. When hematopoietic waves have been specifically inspected, emergence of YS and AGM-like stages were conflated in time (Pearson et al., 2015; Murry and Keller, 2008), suggesting a bias towards intrinsic hematopoietic cell potential, at the expense of cell fate modulation by time-coherent extrinsic physiological cues, affecting the nature of cellular output. The present hGx system, in contrast, successfully deconvolutes YS-like and AGM-like stages of blood production, including capturing differences in the HE generation, suggesting a level of organisation amenable to further development into an integrated hemogenic niche. Although the system is shy of achieving HSC maturation in situ, it can mature in a supportive microenvironment, in this case provided by the adrenal gland (Schyrr et al., 2023), to produce engrafting multilineage progenitors, and potentially HSCs. This suggests that access to the relevant hematopoietic molecular programs is preserved. Incipient specification of non-hematopoietic tissues shown to support HSC emergence, including PDGFRA+ mesenchymal (Chandrakanthan et al., 2022) and neuro-mesenchymal (Miladinovic et al., 2024) cells, and neural-crest derivatives (Kapeni et al., 2022; Fitch et al., 2012), occurs in sync with development of AGM-like hematopoiesis, to indicate the presence of cues that with further maturation may facilitate HSPC specification at the appropriate developmental time.

We have observed that the nature of the symmetry-breaking pulse affects the specification of hemato-endothelial programs, with a more robust Flk1/Kdr activation in the presence of Activin A, potentially balancing mesodermal derivatives and the anterior-posterior axis. The hematopoietic output of cardiogenic gastruloids (Rossi et al., 2022) does not seem to require Activin A, but their blood formation is of lower efficiency, at least at the comparable timepoints of 144 and 168h, and apparently biased towards erythroid formation, reflecting its YS, and potentially primitive, nature. In the absence of additional cytokines after the CHI99021 pulse, standard gastruloid protocols (Beccari et al., 2018) capture a similar transcriptional erythrogenic wave. Although Rossi et al. (2022) reported the formation of spleen colonies, these are difficult to ascertain from the data published, and the engraftment capacity of cardiogenic gastruloid-derived blood remains uncertain. Our hGx system has more robust CD41+ cell production at 144-168h, which may reflect distinct tissue induction by the CHI99021+Activin A pulse. Interestingly, through extending the hGx culture, and in addition to AGM-like hematopoiesis, we also captured the potential to form renal tubular structures upon in vivo maturation, suggesting induction of lateral plate and intermediate mesoderm from which the AGM region derives, and putative reconstitution of the intra-embryonic hematopoietic niche. An additional challenge is tissue organisation: non-hematopoietic gastruloid models benefitted from mechanical, or chemo-mechanical, support of artificial ECM embedding (Veenvliet et al., 2020; van den Brink, Susanne C et al., 2020), to match temporal coherence and tissue polarisation, with detailed tissue architecture illustrated by somite formation. Prior to HSC emergence in the AGM, critical signals such as SCF and Shh, which we have delivered to the cultures, are secreted by different cells on opposite dorsal-ventral locations (Souilhol et al., 2016), indicating the potential relevance of tissue organisation to improve and potentially extend hematopoietic formation. Moreover, HSPC formation in the AGM is facilitated by circulatory shear stress (North et al., 2009; Diaz et al., 2015; Lundin et al., 2020), suggesting that introduction of flow may improve hemato-endothelial organization and boost blood production.

Over the last decade, the effort to achieve robust hematopoietic production from PSC, mouse and human, has been centred on the generation of HSC, a critical translational need (Chang et al., 2023; Dijkhuis et al., 2023; Fidanza and Forrester, 2021). There are good indications that this is achievable, both with (Vo and Daley, 2015; Dang et al., 2002) and without (Piau et al., 2023) genetic manipulation. However, success remains episodic, mainly because mechanistic understanding of processes underpinning blood generation during development is limiting, exacerbated by limiting access to human embryonic and fetal material. On-going optimization of gastruloid models and the success at integration of organoid cultures in microfluidic systems (Saorin, Caligiuri and Rizzolio, 2023; Quintard et al., 2024) will allow us to further optimize hGx culture. In conclusion, the hGx represents a faithful in vitro reference to replace and complement embryo studies, holding promise to understand the biology of healthy and malignant developmental haematopoiesis and advance blood production for clinical use.

Experimental procedures

Materials

Methods

Cell culture

Flk-1-GFP (Jakobsson et al., 2010), Flk-1-GFP::Rosa26-BFP, Sox17-GFP (Niakan et al., 2010), T/Bra::GFP (Fehling et al., 2003), and E14Tg2A (Hooper et al., 1987) mouse embryonic stem cells (mES) lines were cultured in ES+LIF medium in gelatinized (0.1% gelatin) with daily medium change, as previously described (Turner et al., 2017). The ES+LIF medium contained 500 ml Glasgow MEM BHK-21 (Gibco), 50 ml of fetal bovine serum (FBS, Embryonic Stem Cells tested, biosera, Nuaillé, France), 5 ml GlutaMAX supplement (Gibco), 5 ml MEM Non-Essential Amino Acids (Gibco), 5 ml Sodium pyruvate solution (100 mM, Gibco), and 1 ml 2-Mercaptoethanol (50mM, Gibco). Murine LIF (Peprotech) was added at 1000 U/ml. HEK293T cells were grown in Dulbecco’s modified Eagle medium (Gibco) supplemented with 10% FBS (Gibco) and 1% penicillin/streptomycin (Gibco). All cultures were kept at 37°C and 5% CO2.

Hemogenic gastruloids assembly, culture, and dissociation

mES were maintained in ES+LIF medium and transferred to 2i+LIF (containing Chiron and MEK inhibitor PD03) for 24 hours prior to the assembly into gastruloids. 250 untransduced mES cells (or 400 for GX-MNX1 and GX-EV) were seeded in each well of a U-bottom, cell-repellent 96-well plate (Greiner Bio-One, Stonehouse, UK) in 40 μl of N2B27 medium [Takara Bio or home-made as Cold Spring Harbour Protocols ‘N2B27 Medium’ (2017)]. The plate was centrifuged at 750 rpm for 2 minutes to promote deposition and aggregation of the cells and was then incubated at 37°C, 5% CO2 for 48 hours. After 48 hours, 150 μl of N2B27 medium supplemented with 100 ng/ml Activin A (QKine, Cambridge, UK) and 3 μM chiron (Peprotech) was added to each well. At 72 hours, 150 μl of medium were removed, without disrupting the gastruloids in the wells. 100 μl of N2B27 with 5 ng/ml of Vegf and Fgf2 each (Peprotech) were added to each well. From 72 h to 144 h, each day 100 μl of medium were removed and replaced with N2B27 + Vegf + Fgf2. At 144 h, the medium was further supplemented with Shh at 20 ng/ml. From 168 h to 216 h, the medium was N2B27 + 5 ng/ml Vegf, plus 20 ng/ml mTpo, 100 ng/ml mFlt3l, and 100 ng/ml mScf (Peprotech). To dissociate cells from the gastruloid structures, medium was removed and individual gastruloids were collected using a pipette and precipitated at the bottom of a microcentrifuge tube. The remaining medium was aspirated and the bulk of gastruloids was washed in PBS. 50 μl of TrypLE express was added to pelleted gastruloids to be incubated at 37°C for 2 minutes to dissociate cells.

Animals and implantation

Female athymic nude-Foxn1nu (nu/nu) mice (Envigo, Bresso, Italy) were housed under pathogen-free conditions. In accordance with the “3Rs policy”, experiments were reviewed and approved by the Animal Welfare Body (OPBA) of IRCCS Ospedale Policlinico San Martino and by the Italian Ministry of Health (n. 883/2020-PR).

Intact hGx (3 per mouse in 10 μL of PBS) were inoculated in the adrenal gland of five-week-old mice. Briefly, mice were anesthetized with a mixture of xylazine (ROMPUN, Bayer) and ketamine (LOBOTOR, Acme S.r.l.) and injected with hGx, after laparotomy, in the capsule of the left adrenal gland, as previously described (Brignole et al., 2023; Pastorino et al., 2003). All mice survived surgery. Control mice (CTR) received vehicle only. Mice body weight and general physical conditions were recorded daily. One month after hGx inoculation, mice were sacrificed and spleen and bone marrow (BM) from femurs collected. BM cells and spleens, mechanically reduced to cell suspension, were stored at -80°C until use. Adrenal glands were paraffin-embedded for subsequent IHC studies.

Methylcellulose colony forming assays (CFC)

Disassembled gastruloids cells were plated on Mouse Methylcellulose Complete Media (R&D Systems). Cells were first suspended in Iscove’s Modified Dulbecco’s Medium (IMDM, Gibco) with 20% FBS (Gibco) before addition to the methylcellulose medium. Cells were plated in duplicate 35 mm dishes with 2×105 cells/plate. Plates were incubated at 37°C and 5% CO2 for 10 days, when colonies were scored. For serial replating experiments, cells in methylcellulose were collected and washed in phosphate buffer saline (PBS) to achieve single-cell suspensions and replated as described above. Dissociated gastruloid cells from CFC were collected as cell suspensions in PBS and centrifuged onto slides (Shandon Cytospin 2 Cytocentrifuge) at 350rpm for 5-7min, with a PBS-only pre-spin at 1000rpm, 2min. Slides were stained in Giemsa-Wright stain for 30 seconds followed by addition of phosphate buffer pH 6.5 for 5 min.

Immunofluorescence staining

Immunostaining of whole gastruloids was performed as described before (Baillie-Johnson et al., 2015). Briefly, gastruloids were fixed in 4% paraformaldehyde (PFA) dissolved in PBS for 4 hours at 4°C on orbital shaking and permeabilised in PBSFT (10% FBS and 0.2% Triton X-100), followed by one hour blocking in PBSFT at 4°C on orbital shaking. Antibody dilutions were made in PBSFT at 1:200 for primary and 1:500 for secondary antibodies. Antibody incubations were performed overnight at 4°C on orbital shaking, and subjected to optical clearing overnight in ScaleS4 clearing solution. Individual gastruloids were then mounted on glass coverslips by pipetting as droplets in ScaleS4 and DAPI nuclear stain.

Histology and H&E staining

Samples were fixed with 10% formalin at room temperature for 48 hours followed by an ethanol series of ethanol and paraffin; ethanol 70% (90 minutes), ethanol 80% (120 minutes), ethanol 95% (180 minutes), 3 times ethanol 100% of 120, 120 and 180 minutes, 50% ethanol/paraffin (3x120 minutes) and 3 times paraffin (60, 90 and 120 minutes) before embedding in histological cassettes. Blocks were sectioned with a HistoCore BIOCUT Microtome (Leica) at 10 mm and slides were stained with Haematoxylin/Eosin.

Imaging

Images of cultured gastruloids and CFC plates were captured using the Cytation 5 Cell Imaging Multi-Mode Reader (Biotek) plate reader using bright field and FITC channels. ImageJ (Schneider et al., 2012) was used for gastruloid size quantification. Confocal microscopy was performed on LSM700 on a Zeiss Axiovert 200 M with Zeiss EC Plan-Neofluar 10x/0.30 M27 and Zeiss LD Plan-Neofluar 20x/0.4 M27 objective lens. Tissue sections were imaged with a THUNDER Imager 3D microscope system (Leica).

Lentiviral vector packaging and transduction

The lentiviral overexpression vector pWPT-LSSmOrange-PQR was used to clone the MNX1 gene cDNA. The viral packaging vectors pCMV and pMD2.G, described in Pina et al. (2008), were used to assemble lentiviral particles using HEK293T cells via transfection using TurboFect Reagent (Invitrogen). Transduction of mES cells was performed overnight by addition of lentivirus to culture medium and washed the following day (Moris et al., 2018). Transduced cells were sorted for positivity to LSSmOrange.

Flow cytometry

Surface cell marker analysis was performed by staining using the antibodies listed in Key Resources Table. Disassembled gastruloids cells were resuspended in PBS, 2% FBS and 0.5 mM EDTA and stained at a dilution of 1:100 for primary antibodies for 20 minutes at 4°C. When indicated, streptavidin was added at a dilution of 1:200. Analysis was performed on ACEA Novocyte (Agilent) or AttuneNxT (Thermo) analyzers, using the respective software packages. Cell sorting was performed using a CS&T calibrated BDFACS Aria III system (488nm 40 mW, 633nm 20 mW, 405nm 30 mW, and 561nm 50 mW), set with the 100μm nozzle at 20PSI and a 4-way purity mask or, in the case of hGx adrenal implants, a Beckman Coulter CytoFlex SRT (488nm 50 mW, 633nm 100 mW, 405nm 90 mW, and 561nm 30 mW), set with 100μm nozzle at 15PSI, also using 4-way purity. Single-cell deposition in 96-well plates was performed using single-cell sorting mode. Intact cells were gated on FSC-A vs SSC-A plot, followed by doublet exclusion on FSC-A vs FSC-H and SSC-A vs SSC-H, prior to gating on fluorescent parameters for the markers described in the results.

Single cell RNA sequencing of time-resolved hemogenic gastruloids

Gastruloids were collected at different timepoints of the protocol, disassembled and FACS deposited into 96-well plates, either as unsorted (global) or sorted by CD45, CD41 and C-Kit/ScaI markers (sorted). RNA from the cells were extracted from single cells using Smart-seq2 technology at 500Kb and depth of 151 bases. Sequencing reads were quality-checked using FastQC (v0.11.9). We trimmed the samples using trimGalore! (0.6.6) with a cutoff of 30, clipping 15 base pairs and retaining reads of more than 100 bases. Alignment was performed on STAR (2.7.8a) and Mus musculus annotations from GENCODE vM26. Aligned BAM files were annotated using featureCounts (v2.0.1) and count matrices were computed in python by directly accessing the BAM files with gtfparse packages and collapsing lanes.

For quality control, based on the histogram of counts and multimodality distributions, we set a minimum count threshold of 200000 counts, a minimum threshold of 1000 expressed genes, and a maximum threshold of 20% mitochondrial fraction per cell. We performed the same procedure of the second dataset with a more restrictive threshold of 400000 counts and similar expressed genes and mitochondrial thresholds. Cells that did not pass the quality control metrics were omitted from analysis. We normalized the cells to the mean count number per dataset and applied a plus-one-log transformation of the data before proceeding to the downstream analysis.

Dimensionality reduction was performed on feature selection of the gene space using the function scanpy.highly_varying_genes with default parameters. Selection of principal components in principal component analysis (PCA) was performed by heuristic elbow method. Nearest neighbor analysis was constructed by KNN graphs using a correlation metric and 10 nearest neighbors. Data was projected for low dimensional visualization using the UMAP algorithm with default parameters as implemented in scanpy.tl.umap. We used the leiden algorithm as implemented in scanpy.tl.leiden to partition the data into clusters. In order to assess the election of the resolution parameter we used Newman-Girvan modularity as a metric of clustering quality. Differential expressed genes were computed comparing each cluster against the rest using the Wilcoxon test with Benjamini-Hochberg correction.

Annotation and projection to additional datasets

Raw counts matrices from Thambyrajah et al. (2024) were downloaded and processed following the same pipeline as described above for scRNA-seq of gastruloids to generate UMAP and clustering. We implemented the scmap algorithm to compare our scRNA sequencing of gastruloids with the available datasets. We reduced the dimensionality of the space by selecting highly varying genes from the annotated dataset. Then, we constructed a KNN classifier with correlation metric and computed the nearest neighbors of the target data. If neighbors with correlation metrics below 0.7 default standards, the projected cells were not projected onto any cell from the annotated dataset. To visualize the cells over the UMAP plots of the other datasets, we constructed a KNN regressor with three neighbors and a correlation metric. Confusion matrices were used to visualize the overlap between gastruloid clusters and the embryo dataset assigned clusters using scmap.

Bulk RNA sequencing

Total RNA was extracted from disassembled gastruloid cells at 216 hours. Sequencing was performed on NovaSeq PE150 platform, at 20M paired-end reads per sample. Tophat2 with Bowtie2 were used to map paired-end reads to the reference Mus musculus genome build GRCm39. GENCODE Release M30 (Frankish et al., 2019) was used as the reference mouse genome annotation. Aligned reads were filtered by quality using samtools (Li, H. et al., 2009) with a minimum threshold set at 30 (q30). Transcript assembly and quantification was achieved using htseq (Putri et al., 2022). Differential expression between sample and control was performed by collapsing technical replicates for each condition using Deseq2 (Love, Huber and Anders, 2014) in R environment (Deseq2 library v 1.32.0). The differential expression was expressed in the form of log2 fold change and filtered by false discovery rate (FDR) of 0.05.

Gene list enrichment analyses

Gene ontology (GO) analysis was performed in ExpressAnalyst (available at www.expressanalyst.ca) using the PANTHER or GO Biological Process (BP) repository. GO terms and pathways were filtered by false discovery rate (FDR) with a cut-off of ≤ 0.1 for meaningful association. EnrichR (Xie et al., 2021; Kuleshov et al., 2016; Chen, E. Y. et al., 2013) was used for cell type analysis using the Panglao DB (Franzén, Gan and Björkegren, 2019) databases using a FDR threshold of ≤ 0.1 or p value ≤ 0.01, where specified, as well as transcription factor (TF) binding site enrichment using the ENCODE and ChEA Consensus TFs from ChIP database.

Gene Set Enrichment Analysis (GSEA)

Custom gene signatures were used as gene sets for GSEA analysis (Subramanian et al., 2005) on the GSEA software v4.2.3 on RNA sequencing expression values in counts units. GSEA was ran in 10000 permutations on gene set using the weighted Signal2Noise metric. Enrichment metrics are shown as normalized enrichment score (NES) and filtered by FDR ≤ 0.05. Leading edge genes (LEGs) are genes with a “Yes” values for core enrichment. For AML patient analysis, clinical phenotype and expression data (in counts units) were extracted from the GDC TARGET-AML cohorts in the Therapeutically Applicable Research to Generate Effective Treatments project (TARGET, https://ocg.cancer.gov/programs/target), downloaded from the University of California Santa Cruz (UCSC) Xena public repository (last accessed 31st August 2022). Patient samples were selected according to the reported karyotype to include t(7;12), inv(16), MLL, normal karyotype, and t(8;21). GSEA was performed comparing RNA sequencing counts of t(7;12) samples against pooled AML subtypes (inv(16), MLL, normal karyotype and t(8;21)) as “other AML”.

Real-time polymerase chain reaction (qPCR)

Extracted RNA was reverse-transcribed into complementary DNA (cDNA) using High-Capacity RNA-to-cDNA Kit (Applied Biosystems). QPCR was performed using FastGene 2x IC Green Universal qPCR Mix (Nippon Genetics, Duren, Germany) using primers for human MNX1 (forward 5-GTTCAAGCTCAACAAGTACC-3; reverse 5-GGTTCTGGAACCAAATCTTC-3) (Gulino et al., 2021) and Ppia for endogenous control (forward 5-TTACCCATCAAACCATTCCTTCTG-3; reverse 5-AACCCAAAGAACTTCAGTGAGAGC-3) (Moris et al., 2018). Differential gene expression was calculated using the delta delta Ct (ΔΔCt) method.

Polymerase chain reaction (PCR)

Genomic DNA was extracted using Monarch Genomic DNA Purification Kit (New England Biolabs, Hitchin, UK) using manufacturer’s instructions. PCR on genomic DNA was performed using Phire PCR Master Mix (Thermo Scientific) to amplify the BFP locus using forward primer 5’-GCACCGTGGACAACCATCACTT-3’ and reverse primer 5’-CAGTTTGCTAGGGAGGTCGC-3’.

Statistical analysis

Experiments were performed at least in triplicates, unless specified otherwise. Data are plotted to include standard deviation (+/-SD) between replicates. Statistical significance was set at a threshold of p value < 0.05. Statistical analysis was performed in R environment (version 4.1.3) or using GraphPad Prism 8.0 software and is detailed in respective figure legends.

Data availability

Raw data as well as processed count matrices and post-processed files from single-cell RNA-seq for the time-resolved data is available at E-MTAB-12148. Bulk RNA-seq of MNX1 overexpressing gastruloids is available at Array Express with accession code E-MTAB-12173. The post-processing was performed in Python on DockerHub: dsblab/single_cell_analysis:0.5. Scripts are available in https://github.com/dsb-lab/blood_gastruloids and Zenodo (https://doi.org/10.5281/zenodo.7053423). The results published here are partly based upon data generated by the Therapeutically Applicable Research to Generate Effective Treatments (TARGET) (https://ocg.cancer.gov/programs/target) initiative, of the Acute Myeloid Leukemia (AML) cohort GDC TARGET-AML. The data used for this analysis are available at https://portal.gdc.cancer.gov/projects and https://xenabrowser.net/.

eLife Assessment

This valuable study presents a mouse gastruloid model that can be used to generate hematopoietic progenitors as well as leukemic cells. However, in its current form, the manuscript is inadequate because the primary claims are not supported. Overall, the hematopoietic progenitor cells generated in this system need to be better defined.

Reviewer #1 (Public review):

Summary

The authors describe a method for gastruloid formation using mouse embryonic stem cells (mESCs) to study YS and AGM-like hematopoietic differentiation. They characterise the gastruloids during nine days of differentiation using a number of techniques including flow cytometry and single-cell RNA sequencing. They compare their findings to a published data set derived from E10-11.5 mouse AGM. At d9, gastruloids were transplanted under the adrenal gland capsule of immunocompromised mice to look for the development of cells capable of engrafting the mouse bone marrow. The authors then applied the gastruloid protocol to study overexpression of Mnx1 which causes infant AML in humans.

In the introduction, the authors define their interpretation of the different waves of hematopoiesis that occur during development. 'The subsequent wave, known as definitive, produces: first, oligopotent erythro-myeloid progenitors (EMPs) in the YS (E8-E8.5); and later myelo-lymphoid progenitors (MLPs - E9.5-E10), multipotent progenitors (MPPs - E10-E11.5), and hematopoietic stem cells (HSCs - E10.5-E11.5), in the aorta-gonad-mesonephros (AGM) region of the embryo proper.' Herein they designate the yolk sac-derived wave of EMP hematopoiesis as definitive, according to convention, although paradoxically it does not develop from intra-embryonic mesoderm or give rise to HSCs.

General comments

The authors make the following claims in the paper:

(1) The development of a protocol for hemogenic gastruloids (hGx) that recapitulates YS and AGM-like waves of blood from HE.

(2) The protocol recapitulates both YS and EMP-MPP embryonic blood development 'with spatial and temporal accuracy'.

(3) The protocol generates HSC precursors capable of short-term engraftment in an adrenal niche.

(4) Overexpression of MNX1 in hGx transforms YS EMP to 'recapitulate patient transcriptional signatures'.

(5) hGx is a model to study normal and leukaemic embryonic hematopoiesis.

There are major concerns with the manuscript. The statements and claims made by the authors are not supported by the data presented, data is overinterpreted, and the conclusions cannot be justified. Furthermore, the data is presented in a way that makes it difficult for the reader to follow the narrative, causing confusion. The authors have not discussed how their hGx compares to the previously published mouse embryoid body protocols used to model early development and hematopoiesis.

Specific points

(1) It is claimed that HGxs capture cellularity and topography of developmental blood formation. The hGx protocol described in the manuscript is a modification of a previously published gastruloid protocol (Rossi et al 2022). The rationale for the protocol modifications is not fully explained or justified. There is a lack of novelty in the presented protocol as the only modifications appear to be the inclusion of Activin A and an extension of the differentiation period from 7 to 9 days of culture. No direct comparison has been made between the two versions of gastruloid differentiation to justify the changes.

The inclusion of Activin A at high concentration at the beginning of differentiation would be expected to pattern endoderm rather than mesoderm. BMP signaling is required to induce Flk1+ mesoderm, even in the presence of Wnt. FACS analysis of the hGx during differentiation is needed to demonstrate the co-expression of Flk1-GFP and lineage markers such as CD34 to indicate patterning of endothelium from Flk1+ mesoderm. The FACS plots in Figure 1 show c-Kit expression but very little VE-cadherin which suggests that CD34 is not induced. Early endoderm expresses c-Kit, CXCR4, and Epcam but not CD34 which could account for the lack of vascular structures within the hGx as shown in Figure 1E.

(2) The protocol has been incompletely characterised, and the authors have not shown how they can distinguish between either wave of Yolk Sac (YS) hematopoiesis (primitive erythroid/macrophage and erythro-myeloid EMP) or between YS and intraembryonic Aorta-Gonad-Mesonephros (AGM) hematopoiesis. No evidence of germ layer specification has been presented to confirm gastruloid formation, organisation, and functional ability to mimic early development. Furthermore, differentiation of YS primitive and YS EMP stages of development in vitro should result in the efficient generation of CD34+ endothelial and hematopoietic cells. There is no flow cytometry analysis showing the kinetics of CD34 cell generation during differentiation. Benchmarking the hGx against developing mouse YS and embryo data sets would be an important verification.

Single-cell RNA sequencing was used to compare hGx with mouse AGM. The authors incorrectly conclude that ' ..specification of endothelial and HE cells in hGx follows with time-dependent developmental progression into putative AGM-like HE..' And, '...HE-projected hGx cells.......expressed Gata2 but not Runx1, Myb, or Gfi1b..' Hemogenic endothelium is defined by the expression of Runx1 and Gfli1b is downstream of Runx1.

(3) The hGx protocol 'generates hematopoietic SC precursors capable of short-term engraftment' is not supported by the data presented. Short-term engraftment would be confirmed by flow cytometric detection of hematopoietic cells within the recipient bone marrow, spleen, thymus, and peripheral blood that expressed the BFP transgene. This analysis was not provided. PCR detection of transcripts, following an unspecified number of amplification cycles, as shown in Figure 3G (incorrectly referred to as Figure 3F in the legend) is not acceptable evidence for engraftment. Transplanted hGx formed teratoma-like structures, with hematopoietic cells present at the site of transplant only analysed histologically. Indeed, the quality of the images provided does not provide convincing validation that donor-derived hematopoietic cells were present in the grafts.

There is no justification for the authors' conclusion that '... the data suggest that 216h hGx generate AGM-like pre-HSC capable of at least short-term multilineage engraftment upon maturation...'. Indeed, this statement is in conflict with previous studies demonstrating that pre-HSCs in the dorsal aorta of the mouse embryo are immature and actually incapable of engraftment.

The statement '...low-level production of engrafting cells recapitulates their rarity in vivo, in agreement with the embryo-like qualities of the gastruloid system....' is incorrect. Firstly, no evidence has been provided to show the hGx has formed a dorsal aorta facsimile capable of generating cells with engrafting capacity. Secondly, although engrafting cells are rare in the AGM, approximately one per embryo, they are capable of robust and extensive engraftment upon transplantation.

(4) Expression MNX1 transcript and protein in hematopoietic cells in MNX1 rearranged acute myeloid leukaemia (AML) is one cause of AML in infants. In the hGX model of this disease, Mnx1 is overexpressed in the mESCs that are used to form gastruloids. Mnx1 overexpression seems to confer an overall growth advantage on the hGx and increase the serial replating capacity of the small number of hematopoietic cells that are generated. The inefficiency with which the hGx model generates hematopoietic cells makes it difficult to model this disease. The poor quality of the cytospin images prevents accurate identification of cells. The statement that the kit-expressing cells represent leukemic blast cells is not sufficiently validated to support this conclusion. What other stem cell genes are expressed? Surface kit expression also marks mast cells, frequently seen in clonogenic assays of blood cells. Flow cytometric and gene expression analyses using known markers would be required.

(5) In human infant MNX1 AML, the mutation is thought to arise at the fetal liver stage of development. There is no evidence that this developmental stage is mimicked in the hGx model.

Reviewer #2 (Public review):

Summary:

In this manuscript, the authors develop an exciting new hemogenic gastruloid (hGX) system, which they claim reproduces the sequential generation of various blood cell types. The key advantage of this cellular system would be its potential to more accurately recapitulate the spatiotemporal emergence of hematopoietic progenitors within their physiological niche compared to other available in vitro systems. The authors present a large set of data and also validate their new system in the context of investigating infant leukemia.

Strengths:

The development of this new in vitro system for generating hematopoietic cells is innovative and addresses a significant drawback of current in vitro models. The authors present a substantial dataset to characterize this system, and they also validate its application in the context of investigating infant leukemia.

Weaknesses:

The thorough characterization and full demonstration that the cells produced truly represent distinct waves of hematopoietic progenitors are incomplete. The data presented to support the generation of late yolk sac (YS) progenitors, such as lymphoid cells, and aortic-gonad-mesonephros (AGM)-like progenitors, including pre-hematopoietic stem cells (pre-HSCs), by this system are not entirely convincing. Given that this is likely the manuscript's most crucial claim, it warrants further scrutiny and direct experimental validation. Ideally, the identity of these progenitors should be further demonstrated by directly assessing their ability to differentiate into lymphoid cells or fully functional HSCs. Instead, the authors primarily rely on scRNA-seq data and a very limited set of markers (e.g., Ikzf1 and Mllt3) to infer the identity and functionality of these cells. Many of these markers are shared among various types of blood progenitors, and only a well-defined combination of markers could offer some assurance of the lymphoid and pre-HSC nature of these cells, although this would still be limited in the absence of functional assays.

The identification of a pre-HSC-like CD45⁺CD41⁻/lo c-Kit⁺VE-Cadherin⁺ cell population is presented as evidence supporting the generation of pre-HSCs by this system, but this claim is questionable. This FACS profile may also be present in progenitors generated in the yolk sac such as early erythro-myeloid progenitors (EMPs). It is only within the AGM context, and in conjunction with further functional assays demonstrating the ability of these cells to differentiate into HSCs and contribute to long-term repopulation, that this profile could be strongly associated with pre-HSCs. In the absence of such data, the cells exhibiting this profile in the current system cannot be conclusively identified as true pre-HSCs.

The engraftment data presented are also not fully convincing, as the observed repopulation is very limited and evaluated only at 4 weeks post-transplantation. The cells detected after 4 weeks could represent the progeny of EMPs that have been shown to provide transient repopulation rather than true HSCs.

Reviewer #3 (Public review):

In this study, the authors employ a mouse ES-derived "hemogenic gastruloid" model which they generated and which they claim to be able to deconvolute YS and AGM stages of blood production in vitro. This work could represent a valuable resource for the field. However, in general, I find the conclusions in this manuscript poorly supported by the data presented. Importantly, it isn't clear what exactly are the "YS" and the "AGM"-like stages identified in the culture and where is the data that backs up this claim. In my opinion, the data in this manuscript lack convincing evidence that can enable us to identify what kind of hematopoietic progenitor cells are generated in this system. Therefore, the statement that "our study has positioned the MNX1-OE target cell within the YS-EMP stage (line 540)" is not supported by the evidence presented in this study. Overall, the system seems to be very preliminary and requires further optimization before those claims can be made.

Specific comments below:

(1) The flow cytometric analysis of gastruloids presented in Figure 1 C-D is puzzling. There is a large % of c-Kit+ cells generated, but few VE-Cad+ Kit+ double positive cells. Similarly, there are many CD41+ cells, but very few CD45+ cells, which one would expect to appear toward the end of the differentiation process if blood cells are actually generated. It would be useful to present this analysis as consecutive gating (i.e. evaluating CD41 and CD45 within VE-Cad+ Kit+ cells, especially if the authors think that the presence of VE-Cad+ Kit+ cells is suggestive of EHT). The quantification presented in D is misleading as the scale of each graph is different.

(2) The imaging presented in Figure 1E is very unconvincing. C-Kit and CD45 signals appear as speckles and not as membrane/cell surfaces as they should. This experiment should be repeated and nuclear stain (i.e. DAPI) should be included.

(3) Overall, I am not convinced that hematopoietic cells are consistently generated in these organoids. The authors should sort hematopoietic cells and perform May-Grunwald Giemsa stainings as they did in Figure 6 to confirm the nature of the blood cells generated.

(4) The scRNAseq in Figure 2 is very difficult to interpret. Specific points related to this:

- Cluster annotation in Figure 2a is missing and should be included.

- Why do the heatmaps show the expression of genes within sorted cells? Couldn't the authors show expression within clusters of hematopoietic cells as identified transcriptionally (which ones are they? See previous point)? Gene names are illegible.

- I see no expression of Hlf or Myb in CD45+ cells (Figure 2G). Hlf is not expressed by any of the populations examined (panels E, F, G). This suggests no MPP or pre-HSC are generated in the culture, contrary to what is stated in lines 242-245. (PMID 31076455 and 34589491).

Later on, it is again stated that "hGx cells... lacked detection of HSC genes like Hlf, Gfi1, or Hoxa9" (lines 281-283). To me, this is proof of the absence of AGM-like hematopoiesis generated in those gastruloids.

(5) Mapping of scRNA-Seq data onto the dataset by Thambyrajah et al. is not proof of the generation of AGM HE. The dataset they are mapping to only contains AGM cells, therefore cells do not have the option to map onto something that is not AGM. The authors should try mapping to other publicly available datasets also including YS cells.

(6) Conclusions in Figure 3, named "hGx specify cells with preHSC characteristics" are not supported by the data presented here. Again, I am not convinced that hematopoietic cells can be efficiently generated in this system, and certainly not HSCs or pre-HSCs.

- FACS analysis in 3A is again very unconvincing. I do not think the population identified as c-Kit+ CD144+ is real. Also, why not try gating the other way around, as commonly done (e.g. VE-Cad+ Kit+ and then CD41/CD45)?

- The authors must have tried really hard, but the lack of short- or long-engraftment in a number of immunodeficient mouse models (lines 305-313) really suggests that no blood progenitors are generated in their system. I am not familiar with the adrenal gland transplant system, but it seems like a very non-physiological system for trying to assess the maturation of putative pre-HSCs. The data supporting the engraftment of these mice, essentially seen only by PCR and in some cases with a very low threshold for detection, are very weak, and again unconvincing. It is stated that "BFP engraftment of the Spl and BM by flow cytometry was very low level albeit consistently above control (Fig. S4E)" (lines 337-338). I do not think that two dots in a dot plot can be presented as evidence of engraftment.

(7) Given the above, I find that the foundations needed for extracting meaningful data from the system when perturbed are very shaky at best. Nevertheless, the authors proceed to overexpress MNX1 by LV transduction, a system previously shown to transform fetal liver cells, mimicking the effect of the t(7;12) AML-associated translocation. Comments on this section:

- The increase in the size of the organoid when MNX1 is expressed is a very unspecific finding and not necessarily an indication of any hematopoietic effect of MNX1 OE.

- The mild increase of cKit+ cells (Figure 4E) at the 144hr timepoint and the lack of any changes in CD41+ or CD45+ cells suggests that the increase in Kit+ cells % is not due to any hematopoietic effect of MNX1 OE. No hematopoietic GO categories are seen in RNA seq analysis, which supports this interpretation. Could it be that just endothelial cells are being generated?

(8) There seems to be a relatively convincing increase in replating potential upon MNX1-OE, but this experiment has been poorly characterized. What type of colonies are generated? What exactly is the "proportion of colony forming cells" in Figures 5B-D? The colony increase is accompanied by an increase in Kit+ cells; however, the flow cytometry analysis has not been quantified.

(9) Do hGx cells engraft upon MNX1-OE? This experiment, which appears not to have been performed, is essential to conclude that leukemic transformation has occurred.

Author response:

The following is the authors’ response to the original reviews.

Reviewer #1 (Public review):

Summary

The authors describe a method for gastruloid formation using mouse embryonic stem cells (mESCs) to study YS and AGM-like hematopoietic differentiation. They characterise the gastruloids during nine days of differentiation using a number of techniques including flow cytometry and single-cell RNA sequencing. They compare their findings to a published data set derived from E10-11.5 mouse AGM. At d9, gastruloids were transplanted under the adrenal gland capsule of immunocompromised mice to look for the development of cells capable of engrafting the mouse bone marrow. The authors then applied the gastruloid protocol to study overexpression of Mnx1 which causes infant AML in humans.

In the introduction, the authors define their interpretation of the different waves of hematopoiesis that occur during development. 'The subsequent wave, known as definitive, produces: first, oligopotent erythro-myeloid progenitors (EMPs) in the YS (E8-E8.5); and later myelo-lymphoid progenitors (MLPs - E9.5-E10), multipotent progenitors (MPPs - E10-E11.5), and hematopoietic stem cells (HSCs - E10.5-E11.5), in the aorta-gonad-mesonephros (AGM) region of the embryo proper.' Herein they designate the yolk sac-derived wave of EMP hematopoiesis as definitive, according to convention, although paradoxically it does not develop from intra-embryonic mesoderm or give rise to HSCs.

Our definition of primitive and definitive waves is widely used in the field (e.g. PMID: 18204427; PMID: 28299650; PMID: 33681211). Definitive haematopoiesis, encompassing EMP, MLP, MPP and HSC, highlights their origin from haemogenic endothelium, generation of mature cells with adult characteristics from progenitors with multilineage potential and direct and indirect developmental contributions to the intra-embryonic and time-restricted generation of HSCs.

General comments

The authors make the following claims in the paper:

(1) The development of a protocol for hemogenic gastruloids (hGx) that recapitulates YS and AGMlike waves of blood from HE.

(2) The protocol recapitulates both YS and EMP-MPP embryonic blood development 'with spatial and temporal accuracy'.

(3) The protocol generates HSC precursors capable of short-term engraftment in an adrenal niche.

(4) Overexpression of MNX1 in hGx transforms YS EMP to 'recapitulate patient transcriptional signatures'.

(5) hGx is a model to study normal and leukaemic embryonic hematopoiesis.

There are major concerns with the manuscript. The statements and claims made by the authors are not supported by the data presented, data is overinterpreted, and the conclusions cannot be justified. Furthermore, the data is presented in a way that makes it difficult for the reader to follow the narrative, causing confusion. The authors have not discussed how their hGx compares to the previously published mouse embryoid body protocols used to model early development and hematopoiesis. Specific points

(1) It is claimed that HGxs capture cellularity and topography of developmental blood formation. The hGx protocol described in the manuscript is a modification of a previously published gastruloid protocol (Rossi et al 2022). The rationale for the protocol modifications is not fully explained or justified. There is a lack of novelty in the presented protocol as the only modifications appear to be the inclusion of Activin A and an extension of the differentiation period from 7 to 9 days of culture. No direct comparison has been made between the two versions of gastruloid differentiation to justify the changes.

The Reviewer paradoxically claims that the protocol is not novel and that it differs from a previous publication in at least 2 ways – the patterning pulse and the length of the protocol. Of these, the patterning pulse is key. As documented in Fig. 1S1, we cannot obtain Flk1-GFP expression in the absence of Activin A (Fig. 1S1A), and the concentration of Activin A scales activity of the Flk1 locus (Fig. 1S1B). Expression of Flk1 is a fundamental step in haemato-endothelial specification and, accordingly, we do not see CD41 or CD45+ cells in the absence of Activin A. Furthermore, these markers also titrate with the dose of Activin A (in Fig. 1S1B).

Also, in our hands, there is a clear time-dependent progression of marker expression, with sequential acquisition of CD41 and CD45, with the latter not detectable until 192h (Fig. 1C-D), another key difference relative to the Rossi et al (2022) protocol. We suggest, and present further evidence for in this rebuttal and the revised manuscript, that the 192h-timepoint captures the onset of AGM-like haematopoiesis. We have edited the manuscript to clarify the differences and novelty in our protocol (lines 132-143) and provided a more detailed comparison with the report from Rossi et al. (2022) in the Discussion (lines 574-586).

The inclusion of Activin A at high concentration at the beginning of differentiation would be expected to pattern endoderm rather than mesoderm. BMP signaling is required to induce Flk1+ mesoderm, even in the presence of Wnt.

Again, we call the Reviewer’s attention to Fig. 1S1A which clearly shows that Activin A (with no BMP added) is required for induction of Flk1 expression, in the presence of Wnt. Activin A in combination with Wnt, is used in other protocols of haemato-endothelial differentiation from pluripotent cells, with no BMP added in the same step of patterning and differentiation (PMID: 39227582; PMID: 39223325). In the latter protocol, we also call the Reviewer’s attention to the fact that a higher concentration of Activin A precludes the need for BMP4 addition. Finally, one of us has recently reported that Activin A, on its own, will induce Flk1, as well as other anterior mesodermal progenitors (https://www.biorxiv.org/content/10.1101/2025.01.11.632562v1). In addressing the Reviewer’s concerns with the dose of Activin A used, we titrated its concentration against activation of Flk1, confirming optimal Flk1-GFP expression at the 100ng/ml dose used in the manuscript. We have included this data in the manuscript in Figure 1S1B.

FACS analysis of the hGx during differentiation is needed to demonstrate the co-expression of Flk1GFP and lineage markers such as CD34 to indicate patterning of endothelium from Flk1+ mesoderm. The FACS plots in Fig. 1 show C-Kit expression but very little VE-cadherin which suggests that CD34 is not induced. Early endoderm expresses C-Kit, CXCR4, and Epcam, but not CD34 which could account for the lack of vascular structures within the hGx as shown in Fig. 1E.

We were surprised by the Reviewer’s comment that there are no endothelial structures in our haemogenic gastruloids. The presence of a Flk1-GFP+ network is visible in the GFP images in Fig. 1B, from 144h onwards, and is detailed in the revised Fig. 2A, which shows overlap between Flk1GFP and the endothelial marker CD31. In addition, our single-cell RNA-seq data, included in the manuscript, confirms the presence of endothelial cells with a developing endothelial, including arterial, programme. This is now presented in the revised Fig. 3B-D of the manuscript, which updates a representation in the original manuscript. In contrast with the Reviewer’s claims that no endothelial cells are formed, the data show that Kdr (Flk1)+ cells co-express Cdh5/VE-Cadherin and indeed Cd34, attesting to the presence of an endothelial programme. Arterial markers Efnb2, Flt1, and Dll4 are present. A full-blown programme, which also includes haemogenic markers including Sox17, Esam, Cd44 and Mecom is clear at early (144h) and, particularly at late (192h) timepoints in cells sorted on detection of surface C-Kit (Fig. 3B-E in the manuscript). To address the specific point by the Reviewer, we also document co-expression of Flk1-GFP, CD34 and/or CD31 by flow cytometry (Fig. 2S1A-B in the revised manuscript).

To summarise new and revised data in the manuscript in relation to this point:

Immunofluorescence staining showing the Flk1-GFP-defined vascular network in Figure 1E and co-expression of endothelial marker CD31 in Figure 2A. In text: lines 159-163; 178-180.

Flow cytometry analysis of co-expression of Flk1-GFP with CD31 and CD34 in Figure 2S1AD, including controls. In text: 180-187.

Real-time quantitative (q)PCR analysis showing time-dependent expression of haematoendothelial and arterial markers in Figure 2F (specifically Dll4 and Mecom). In text: 200-209.

An improved representation of our scRNA-seq data highlighting key haemato-endothelial markers in Figure 3B-D. In text: 268-304

(2) The protocol has been incompletely characterised, and the authors have not shown how they can distinguish between either wave of Yolk Sac (YS) hematopoiesis (primitive erythroid/macrophage and erythro-myeloid EMP) or between YS and intraembryonic Aorta-Gonad-Mesonephros (AGM) hematopoiesis. No evidence of germ layer specification has been presented to confirm gastruloid formation, organisation, and functional ability to mimic early development. Furthermore, differentiation of YS primitive and YS EMP stages of development in vitro should result in the efficient generation of CD34+ endothelial and hematopoietic cells. There is no flow cytometry analysis showing the kinetics of CD34 cell generation during differentiation. Benchmarking the hGx against developing mouse YS and embryo data sets would be an important verification.

The Reviewer is correct that we have not provided detailed characterisation of the different germ layers, as this was not the focus of the study. In that context, we were surprised by the earlier comment assuming co-expression of C-Kit, Cxcr4 and Epcam, which we did not show, while overlooking the endothelial programme reiterated above, which we have presented. Given our focus on haemato-endothelial specification, we have started the single-cell RNA-seq characterisation of the haemogenic gastruloid at 120h and have not looked specifically at earlier timepoints of embryo patterning. This said, we show the presence of neuroectodermal cells in cluster 9; on the other hand, cluster 7 includes hepatoblast-like cells, denoting endodermal specification (Supplementary File S2). However, in the absence of earlier timepoints and given the bias towards mesodermal specification, we expect that specification of ectodermal and endodermal programmes may be incomplete.

In respect of the contention regarding the capture of YS-like and AGM-like haematopoiesis, we had presented evidence in the original version of the manuscript that haemogenic cells generated during gastruloid differentiation, particularly at late 192h and 216h timepoints project onto highly purified CKit+ CD31+ Gfi1-expressing cells from mouse AGM (PMID: 38383534), providing support for at least partial recapitulation of the corresponding developmental stage. These projections are represented in Fig. 4A, right and 4S1C of the revised manuscript. In distinguishing between YS-like and AGM-like haematopoiesis, we call the Reviewer’s attention to the replotting of the single-cell RNA-seq data already in the manuscript, which we provided in response to point 1 (Fig. 3B-D and 3S2B), which highlights an increase in Sox17, but not Sox18, expression in the 192h haemogenic endothelium, which suggests an association with AGM haematopoiesis (PMID: 20228271). A significant association of Cd44 and Procr expression with the same time-point (Fig. 3B-D in the manuscript), further supports an AGM-like endothelial-to-haematopoietic transition at the 192h timepoint. We have re-analysed the scRNA-seq data to better represent the expression of these markers in Fig. 3A-E and S32B. We agree that it remains challenging to identify markers exclusive to AGM haematopoiesis, which is operationally equated with generation of transplantable haematopoietic stem cells. While HSC generation is a key event characteristic of the AGM, not all AGM haematopoiesis corresponds to HSCs, an important point in evaluating the data presented in the manuscript, and one that is acknowledged by us. The main text has been edited to clarify the experiments pertaining to distinguishing AGM and YS haematopoiesis, which are detailed in lines 180-187, 200-221, 268-304, and 315-356.

Following on the Reviewer’s comments about Cd34, we also inspected co-expression of Cd34 with Cd41 and Cd45, the latter co-expression present in, although not necessarily exclusive to, AGM haematopoiesis. Reassuringly, we observed clear co-expression with both markers (Author response image 1), in addition to a CD41+CD34- population, which likely reflects YS EMP-independent erythropoiesis. Flow cytometry analysis of co-expression of CD31 and CD34 in CD41+ and CD45+ populations at 144h and 216h timepoints has been included in Fig. 2B-D, Fig. 2S1A-D, including controls. In text: 180-187. We have earlier on in the rebuttal highlighted the fact that marker expression is responsive to the levels of Activin A used in the patterning pulse, with the 100ng/ml Activin A used in our protocol superior to 75ng/ml.

Association of CD34 with CD41 and CD45 expression is Activin A-responsive and supports the presence of definitive haematopoiesis.

A. Flow cytometry analysis of CD34 and CD41 expression in 216h-haemogenic gastruloids; two doses of Activin A were used in the patterning pulse with CHI99021 between 48-72h. FMO controls shown. B. Flow cytometry analysis of CD34 and CD45 at 216h in the same experimental conditions.

Given the centrality of this point in comments by all the Reviewers, we have conducted projections of our single-cell RNA-seq data against two studies which (1) capture arterial and haemogenic specification in the para-splanchnopleura (pSP) and AGM region between E8.0 and E11 (Hou et al, PMID: 32203131), and (2) uniquely capture YS, AGM and FL progenitors and the AGM endothelial-tohaematopoietic transition (EHT) in the same scRNA-seq dataset (Zhu et al, PMID: 32392346). Focusing the analysis on the subsets of haemogenic gastruloid cells sorted as CD41+ (144h) C-Kit+ (144h and 192h) and CD45+ (192h and 216h) (now represented in Fig. 3A, and projected onto the studies in Fig. 4A), we show:

(1) That a subset of haemato-endothelial cells from haemogenic gastruloids at 144h to 216h project onto intra-embryonic cells spanning E8.25 to E10 (revised Fig. 4A left and 4S1A). This is in agreement with our original interpretation that 216h are no later than the MPP/pre-HSC state of embryonic development, requiring further maturation to generate engrafting progenitors. We have nevertheless removed specific references to pre-HSC, and instead referred to HSPC/progenitors.

(2) That haemogenic gastruloids contain YS-like (including EMP-like) and AGM-like haematopoietic cells (Fig. 4A centre and 4 S1B). Significantly, some of the cells, particularly CKit-sorted cells with a candidate endothelial and HE-like signature project onto AGM pre-HE and HE, as well as IAHC. Some 144h CD41+ and 192h CD45+ cells also project onto IAHC, suggesting that YS-like and AGM-like programmes arise independently and with partial timedependent organisation in the haemogenic gastruloid model. Later, predominantly 216h cells, have characteristics of MPP/LMPP-like cells from the FL, suggesting a progenitor wave of differentiation.

Altogether, the data support the notion that haemogenic gastruloids capture YS and AGM haematopoiesis until E10, as suggested by us in the manuscript.This re-analysis of the scRNA-seq data which was indeed prompted by challenging and insightful comments from the Reviewers, has been incorporated in the manuscript as described above and further listed here:

Re-clustering and highlights of specific markers in our scRNA-seq data in Figure 3A-E. In text: 268-304.

Projections to mouse embryo datasets in Figure 4A (Figure 4S1A-C; Supplementary File 3). In text: 315-356.

Single-cell RNA sequencing was used to compare hGx with mouse AGM. The authors incorrectly conclude that ' ..specification of endothelial and HE cells in hGx follows with time-dependent developmental progression into putative AGM-like HE..' And, '...HE-projected hGx cells.......expressed Gata2 but not Runx1, Myb, or Gfi1b..' Hemogenic endothelium is defined by the expression of Runx1 and Gfli1b is downstream of Runx1.

As a hierarchy of regulation, Gata2 precedes and drives Runx1 expression at the specification of HE (PMID: 17823307; PMID: 24297996), while Runx1 drives the EHT, upstream of Gfi1b in haematopoietic clusters (PMID: 34517413). Please note that the text segment the Reviewer refers to has been removed from the manuscript, as the analysis is no longer solely focused on projection to Thambyrajah et al (2024) data, and instead gained significantly from the projections on to the Hou et al (2020) and Zhu et al (2020) studies, as detailed above.

(3) The hGx protocol 'generates hematopoietic SC precursors capable of short-term engraftment' is not supported by the data presented. Short-term engraftment would be confirmed by flow cytometric detection of hematopoietic cells within the recipient bone marrow, spleen, thymus, and peripheral blood that expressed the BFP transgene. This analysis was not provided. PCR detection of transcripts, following an unspecified number of amplification cycles, as shown in Figure 3G (incorrectly referred to as Figure 3F in the legend) is not acceptable evidence for engraftment.

We provide the full flow cytometry analysis of spleen engraftment in the 5 mice which received implantation of 216h-haemogenic gastruloids in the adrenal gland and were analysed at 4 weeks; an additional (control) animal received adrenal injection of PBS (Fig. 4B-D in the revised manuscript). In this experiment, the bone marrow collection was limiting, and material was prioritised for PCR (Fig. 4C and full gels in 4S2C in the revised manuscript).

We had previously provided only representative plots of flow cytometry analysis of bone marrow and spleen, which we described as low-level engraftment and were chosen conservatively. The analysis was meant to complement the genomic DNA PCR, where detection was present in only some of the replicates tested per animal. On this note, we confirm that PCR analysis used conventional 40 cycles; the sensitivity had already been shown in the earlier version of the manuscript and is again represented in Fig. 4S2B. We argue that the low level of cytometric and molecular engraftment at 4 weeks, from haemogenic gastruloid-derived progenitors that have not progressed beyond a stage equivalent to E10 (Fig. 4A and Supplementary File 3 in the revised manuscript from scRNAseq projections), and that we have described as requiring additional maturation in vivo, are not surprising. Indeed, as previously shown and now repeated in in Fig. 2B-E (controls in Fig. 2S1E-G) in the revised manuscript, no more than 7 CD45+CD144+ multipotent cells are present per haemogenic gastruloid. We are only able to implant 3 haemogenic gastruloids in the adrenal gland of each transplanted animal.

We have rephrased Results and Discussion in lines 359-415 and 588-621, respectively, to rectify the nature of the engraftment, which we now attribute more generically to progenitors, also in light of the developmental time we could capture in the gastruloids prior to implantation.

Transplanted hGx formed teratoma-like structures, with hematopoietic cells present at the site of transplant only analysed histologically. Indeed, the quality of the images provided does not provide convincing validation that donor-derived hematopoietic cells were present in the grafts.

As stated in the text, the images mean to illustrate that the haemogenic gastruloids developed in situ. Further analysis motivated by the Reviewers’ comments and indeed a subsequent experiment with analysis of engraftment at a later timepoint of 8 weeks (revised Fig. 4E and 4 S2F-G) did not show a direct correspondence between engraftment and in vivo development or expansion, although this occurs in some cases. To be clearer, the observation of donor-derived blood cells in the implanted haemogenic gastruloids would not correspond to engraftment, as we have amply demonstrated that they have generated blood cells in vitro. There is no evidence that there are remaining pluripotent cells in the haemogenic gastruloid after 9 days of differentiation, and it is therefore not clear that the structures observed are teratomas. We specifically comment on this point in the revised manuscript – lines 601-607.

There is no justification for the authors' conclusion that '... the data suggest that 216h hGx generate AGM-like pre-HSC capable of at least short-term multilineage engraftment upon maturation...'. Indeed, this statement is in conflict with previous studies demonstrating that pre-HSCs in the dorsal aorta of the mouse embryo are immature and actually incapable of engraftment.

We have clearly stated that we do not see haematopoietic engraftment through transplantation of dissociated haemogenic gastruloids, which reach the E10 state containing pre-HSC (revised Fig 4A, 4S1A and Supplementary File 3). Instead, we observed rare myelo-erythroid (revised Fig. 4S2F-G) and myelo-lymphoid (revised Fig. 4E) engraftment upon in vivo maturation of haemogenic gastruloids with preserved 3D organisation. These statements are not contradictory. Nevertheless, we have now more cautiously attributed engraftment to the present of progenitors as a generic designation, and not to pre-HSC (lines 412-414 and 588-592 in the revised manuscript).

The statement '...low-level production of engrafting cells recapitulates their rarity in vivo, in agreement with the embryo-like qualities of the gastruloid system....' is incorrect. Firstly, no evidence has been provided to show the hGx has formed a dorsal aorta facsimile capable of generating cells with engrafting capacity. Secondly, although engrafting cells are rare in the AGM, approximately one per embryo, they are capable of robust and extensive engraftment upon transplantation.

As indicated above, the statement in lines 412-414 now reads “Engraftment is erythromyeloid at 4 weeks and lympho-myeloid at 8 weeks, reflecting different classes of progenitors, putatively of YS-like and AGM-like affiliation.” To be clear, with our original statement we meant to highlight that the production of definitive AGM-like haematopoietic progenitors (not all of which are engrafting) in haemogenic gastruloids does not correspond to non-physiological single-lineage programming. We did and do not claim that we achieved production of HSC, which would be long-term engrafting.

(4) Expression MNX1 transcript and protein in hematopoietic cells in MNX1 rearranged acute myeloid leukaemia (AML) is one cause of AML in infants. In the hGX model of this disease, Mnx1 is overexpressed in the mESCs that are used to form gastruloids. Mnx1 overexpression seems to confer an overall growth advantage on the hGx and increase the serial replating capacity of the small number of hematopoietic cells that are generated. The inefficiency with which the hGx model generates hematopoietic cells makes it difficult to model this disease. The poor quality of the cytospin images prevents accurate identification of cells. The statement that the kit-expressing cells represent leukemic blast cells is not sufficiently validated to support this conclusion. What other stem cell genes are expressed? Surface kit expression also marks mast cells, frequently seen in clonogenic assays of blood cells. Flow cytometric and gene expression analyses using known markers would be required.

The haemogenic gastruloid model generates haematopoietic and haemato-endothelial cells. MNX1 expands C-Kit+ cells at 144h, which we show to have a haemato-endothelial signature (see revised Fig. 3A-E, Supplementary File 2). We have added additional flow cytometry data showing that the replating cells from MNX1 express CD31 (Figure 6S1A-B).

Serial replating of CFC assays is a conventional in vitro assay of leukaemia transformation. Critically, colony replating is not maintained in EV control cells, attesting to the transformation potential of MNX1. Although we have not fully-traced the cellular hierarchy of MNX1-driven transformation in the haemogenic gastruloid system, the in vitro replating expands a C-Kit+ cell (revised Fig. 6E), which reflects the surface phenotype of the leukaemia, also recapitulated in the mouse model initiated by MNX1-overexpressing FL cells. Importantly, it recapitulates the transcriptional profile of MNX1leukaemia patients (revised Fig. 7C), which is uniquely expressed by MNX1144h and replated colony cells, but not to MNX1 216h gastruloid cells, arguing against a generic signature of MNX1 overexpression (revised Fig. 7B). Importantly, the MNX1-transformation of haemogenic gastruloid cells is superior to the FL leukaemia model at capturing the unique transcriptional features of MNX1-driven leukaemia, distinct from other forms of AML in the same age group (Fig 7 S1D-F). It is possible that this corresponds to a pre-leukaemia event, and we will explore this in future studies, which are beyond the proof-of-principle nature of this paper.

(5) In human infant MNX1 AML, the mutation is thought to arise at the fetal liver stage of development. There is no evidence that this developmental stage is mimicked in the hGx model.

We never claim that the haemogenic gastruloid model mimics the foetal liver. We propose that susceptibility to MNX1 is at the HE-to-EMP transition. Moreover, and importantly, contrary to the Reviewer’s statement, there is no evidence in the literature that the mutation arises in the foetal liver stage, just that the mutation arises before birth (PMID: 38806630), which is different. In a mouse model of MNX1 overexpression, the authors achieve leukaemia engraftment upon MNX1 overexpression in foetal liver, but not in bone marrow cells (PMID: 37317878). This is in agreement with a vulnerability of embryonic / foetal, but not adult cells to the MNX1 expression caused by the translocation. However, haematopoietic cells in the foetal liver originate from YS and AGM precursors, so the origin of the MNX1susceptible cells can be in those locations, rather than the foetal liver itself.

Reviewer #2 (Public review):

Summary:

In this manuscript, the authors develop an exciting new hemogenic gastruloid (hGX) system, which they claim reproduces the sequential generation of various blood cell types. The key advantage of this cellular system would be its potential to more accurately recapitulate the spatiotemporal emergence of hematopoietic progenitors within their physiological niche compared to other available in vitro systems. The authors present a large set of data and also validate their new system in the context of investigating infant leukemia.

Strengths:

The development of this new in vitro system for generating hematopoietic cells is innovative and addresses a significant drawback of current in vitro models. The authors present a substantial dataset to characterize this system, and they also validate its application in the context of investigating infant leukemia.

Weaknesses:

The thorough characterization and full demonstration that the cells produced truly represent distinct waves of hematopoietic progenitors are incomplete. The data presented to support the generation of late yolk sac (YS) progenitors, such as lymphoid cells, and aortic-gonad-mesonephros (AGM)-like progenitors, including pre-hematopoietic stem cells (pre-HSCs), by this system are not entirely convincing. Given that this is likely the manuscript's most crucial claim, it warrants further scrutiny and direct experimental validation. Ideally, the identity of these progenitors should be further demonstrated by directly assessing their ability to differentiate into lymphoid cells or fully functional HSCs. Instead, the authors primarily rely on scRNA-seq data and a very limited set of markers (e.g., Ikzf1 and Mllt3) to infer the identity and functionality of these cells. Many of these markers are shared among various types of blood progenitors, and only a well-defined combination of markers could offer some assurance of the lymphoid and pre-HSC nature of these cells, although this would still be limited in the absence of functional assays.

The identification of a pre-HSC-like CD45⁺CD41⁻/lo C-Kit⁺VE-Cadherin⁺ cell population is presented as evidence supporting the generation of pre-HSCs by this system, but this claim is questionable. This FACS profile may also be present in progenitors generated in the yolk sac such as early erythromyeloid progenitors (EMPs). It is only within the AGM context, and in conjunction with further functional assays demonstrating the ability of these cells to differentiate into HSCs and contribute to long-term repopulation, that this profile could be strongly associated with pre-HSCs. In the absence of such data, the cells exhibiting this profile in the current system cannot be conclusively identified as true pre-HSCs.

We present 2 additional pieces of evidence to support our claims that we capture YS and AGM stages of haematopoietic development.

(I) In the new Figures 4A and 4 S1A-C and Supplementary File 3 in the revised manuscript, we project our single-cell RNA-seq data onto (1) developing intra-embryonic pSP and AGM between E8 and E11 (Fig. 4A left, 4S1A) and (2) a single-cell RNA-seq study of HE development which combines haemogenic and haematopoietic cells from the YS, the developing HE and IAHC in the AGM, and FL (Fig. 4A centre, 4S1B). Our data maps E8.25-E10, and captures YS EMP and erythroid and myeloid progenitors, as well as AGM pre-HE, HE and IAHC, with some cells matching HSPC and LMPP, as suggested by the projection onto the Thambyrajah et al data set (already presented in the previous version of the manuscript, and now in Fig. 4A right and 4 S1C). The projection of the scRNA-seq data in presented in lines 314-355 of the revised manuscript. The scRNA-seq data itself was refocused on haemato-endothelial programmes as presented in the revised Fig. 3A-E, described in lines 267-303.

(II) Given the difficulty in finding markers that specifically associate with AGM haematopoiesis, we inspected the possibility of capturing different regulatory requirements at different stages of gastruloid development mirroring differential effects in the embryo. Polycomb EZH2 is specifically required for EMP differentiation in the YS, but does not affect AGM-derived haematopoiesis; it is also not required for primitive erythroid cells (PMID: 29555646; PMID: 34857757). We treated haemogenic gastruloids from 120h onwards with either DMSO (0.05%) or GSK126 (0.5uM), and inspected the cellularity of gastruloids at 144h, which we equate with YS-EMP, and 216h – putatively AGM haematopoiesis. We show that EZH2 inhibition / GSK126 treatment specifically reduces %CD41+ cells at 144h, but does not reduce %CD41+ or %CD45+ cells at 216h. We have included this experiment in the manuscript in Fig. 2 S2B-C (in text: 209-221).

These data, together with the scRNA-seq projections described, provide evidence to our claim that 144h haemogenic gastruloids capture YS EMPs, while CD41+ and CD45+ cells isolated at 216h reflect AGM progenitors. We cannot conclude as to the functional nature of the AGM cells from this experiment. The main text has been edited to clarify the experiments pertaining to distinguishing AGM and YS haematopoiesis (lines 180-187; 200-221; 268-304; 315-356).

The engraftment data presented are also not fully convincing, as the observed repopulation is very limited and evaluated only at 4 weeks post-transplantation. The cells detected after 4 weeks could represent the progeny of EMPs that have been shown to provide transient repopulation rather than true HSCs.

In the original version of the manuscript, we stated that there is low level engraftment and did not claim to have generated HSC. Instead, we described cells with short-term engraftment potential. We agree with the Reviewer that the cells we show in the manuscript at 4 weeks could be EMPs (revised Fig. 4B-E and 4 S2D-G). Additionally, we now have 8-week analysis of implant recipients, in which we observed, again low-level, a multi-lineage engraftment of the recipient bone marrow in 1:3 recipients (revised Fig. 4B-E and 4S2F-H). This engraftment is myeloid-lymphoid and therefore likely to have originated in a later progenitor. To be clear, we do not claim that this corresponds to the presence of HSC. It nevertheless supports the maturation of progenitors with engraftment potential. Limiting amounts of material was prioritised for flow cytometry stainings, not allowing PCR analysis. We rephrased Results and Discussion in lines 359-414 and 588-621, respectively, to rectify the nature of the engraftment.

Reviewer #3 (Public review):

In this study, the authors employ a mouse ES-derived "hemogenic gastruloid" model which they generated and which they claim to be able to deconvolute YS and AGM stages of blood production in vitro. This work could represent a valuable resource for the field. However, in general, I find the conclusions in this manuscript poorly supported by the data presented. Importantly, it isn't clear what exactly are the "YS" and the "AGM"-like stages identified in the culture and where is the data that backs up this claim. In my opinion, the data in this manuscript lack convincing evidence that can enable us to identify what kind of hematopoietic progenitor cells are generated in this system. Therefore, the statement that "our study has positioned the MNX1-OE target cell within the YS-EMP stage (line 540)" is not supported by the evidence presented in this study. Overall, the system seems to be very preliminary and requires further optimization before those claims can be made.

Specific comments below:

(1) The flow cytometric analysis of gastruloids presented in Figure 1 C-D is puzzling. There is a large % of C-Kit+ cells generated, but few VE-Cad+ Kit+ double positive cells. Similarly, there are many CD41+ cells, but very few CD45+ cells, which one would expect to appear toward the end of the differentiation process if blood cells are actually generated. It would be useful to present this analysis as consecutive gating (i.e. evaluating CD41 and CD45 within VE-Cad+ Kit+ cells, especially if the authors think that the presence of VE-Cad+ Kit+ cells is suggestive of EHT). The quantification presented in D is misleading as the scale of each graph is different.

Fig. 1C-D provide an overview of haemogenic markers during the timecourse of haemogenic gastruloid differentiation, and does indeed show a late up-regulation of CD45, as the Reviewer points out would be expected. The %CD45+ cells is indeed low. However, we should point out that the haemogenic gastruloid protocol, although biased towards mesodermal outputs, does not aim to achieve pure haematopoietic specification, but rather place it in its embryo-like context. We refute that the scale is misleading: it is a necessity to represent the data in a way that is interpretable by the reader: and we made sure from the outset that the gates (in C) are truly representative and annotated, as are the plot axes (in D). Consecutive gating at the 216h-timepoint is shown and quantified in Fig. 2S1D-F, or in the alternative consecutive gating suggested by the Reviewer, in Author response iamge 2 below. At the request of Reviewer 1, we also analysed CD31 and CD34 within CD41 and CD45 populations, again as validation of the emergent haematopoietic character of the cells obtained. This new analysis is shown in revised Fig. 2B, quantified in 2C.

Flow cytometry analysis of VE-cadherin+ cells in haemogenic gastruloids at 216h of the differentiation protocol, probing co-expression of CD45, CD41 and C-Kit.

(2) The imaging presented in Figure 1E is very unconvincing. C-Kit and CD45 signals appear as speckles and not as membrane/cell surfaces as they should. This experiment should be repeated and nuclear stain (i.e. DAPI) should be included.

We included the requested immunofluorescence staining in Figure 1E (216h). We also show the earlier timepoint of 192h here as Author response image 3. In text: lines 158-162.

Confocal images of haematopoietic production in haemogenic gastruloids.

Wholemount, cleared haemogenic gastruloids were stained for CD45 (pseudo-coloured red) and C-Kit antigens (pseudo-coloured yellow) with indirect staining, as described in the manuscript. Flk1-GFP signal is shown in green. Nuclei are contrasted with DAPI. (A) 192h. (B) 216h.

(3) Overall, I am not convinced that hematopoietic cells are consistently generated in these organoids. The authors should sort hematopoietic cells and perform May-Grunwald Giemsa stainings as they did in Figure 6 to confirm the nature of the blood cells generated.

It is factual that the data are reproducible and complemented by functional assays shown in revised Fig. 2D-E, which clearly demonstrate haematopoietic output. The single-cell RNA-seq data also show expression of a haematopoietic programme, which we have complemented with biologically independent qRT-PCR analysis of the expression of key endothelial and haematopoietic marker and regulatory genes (revised Fig. 2F; in text: 200-209). As requested, we include Giemsa-Wright’s stained cytospins obtained at 216h to illustrate haematopoietic output. These are shown in revised Fig. 2S2A, in text: lines 194-199. Inevitably, the cytospins will be inconclusive as to the presence of endothelial-tohaematopoietic transition or the generation of haematopoietic stem/progenitor cells, as these cells do not have a distinctive morphology.

(4) The scRNAseq in Figure 2 is very difficult to interpret. Specific points related to this: - Cluster annotation in Figure 2a is missing and should be included.

Why do the heatmaps show the expression of genes within sorted cells? Couldn't the authors show expression within clusters of hematopoietic cells as identified transcriptionally (which ones are they? See previous point)? Gene names are illegible.

I see no expression of Hlf or Myb in CD45+ cells (Figure 2G). Hlf is not expressed by any of the populations examined (panels E, F, G). This suggests no MPP or pre-HSC are generated in the culture, contrary to what is stated in lines 242-245. (PMID 31076455 and 34589491).Later on, it is again stated that "hGx cells... lacked detection of HSC genes like Hlf, Gfi1, or Hoxa9" (lines 281-283). To me, this is proof of the absence of AGM-like hematopoiesis generated in those gastruloids.

For a combination of logistic and technical reasons, we performed single-cell RNA-seq using the Smart-Seq2 platform, which is inherently low throughput. We overcame the issue of cell coverage by complementing whole-gastruloid transcriptional profiling at successive time-points with sorting of subpopulations of cells based on individual markers documented in Fig. 1. We clearly stated which platform was used as well as the number and type of cells profiled (Fig. 3S1 and lines 226-241 of the revised manuscript), and our approach is standard. Following suggestions of the Reviewers to further focus our analysis on the haemogenic cellular differentiation within the gastruloids, we revised the presentation of the scRNA-seq data to now provide UMAP projections with representation and quantification of individual genes, including the ones queried by the Reviewer in Fig. 3 and respective supplements. Specifically, re-clustering and highlighting of specific markers are shown in Figure 3A-D and presented in lines 267-303 of the revised manuscript. Complementary independent real-time quantitative (q)PCR analysis showing time-dependent expression of endothelial and haematopoietic markers is now in Figure 2F. In text: 200-208.

(5) Mapping of scRNA-Seq data onto the dataset by Thambyrajah et al. is not proof of the generation of AGM HE. The dataset they are mapping to only contains AGM cells, therefore cells do not have the option to map onto something that is not AGM. The authors should try mapping to other publicly available datasets also including YS cells.

We have done this and the data are presented in Figure 4A (Figure 4S1A) and Supplementary File. In text: 314-355. As detailed in response to Reviewer 1, we have conducted projections of our single-cell RNA-seq data against two studies which (1) capture arterial and haemogenic specification in the para-splanchnopleura (pSP) and AGM region between E8.0 and E11 (Hou et al, PMID: 32203131) (revised Fig. 4A and 4 S1A), and (2) uniquely capture YS, AGM and FL progenitors and the AGM endothelial-to-haematopoietic transition (EHT) in the same scRNA-seq dataset (Zhu et al, PMID: 32392346) (revised Fig. 4A and 4 S1B). Specifically in answering the Reviewers’ point, we show that different subsets of haemogenic gastruloid cells sorted on haemogenic surface markers C-Kit, CD41 and CD45 cluster onto pre-HE and HE, intra-aortic clusters and FL progenitor compartments, and to YS EMP and erythroid and myeloid progenitors. This lends support to our claim that the haemogenic gastruloid system specifies both YS-like and AGM-like cells. Please note that we now do point out that some CD41+ cells at 144h project onto IAC, as do cells at the later timepoints, suggesting that AGM-like and YS-EMP-like waves may overlap at the 144h timepoint (lines…). In the future, we will address specific location of these cells, but that corresponds to a largescale spatial transcriptomics analysis requiring extensive optimisation for section capture which is beyond the scope of this manuscript and this revision.

(6) Conclusions in Figure 3, named "hGx specify cells with preHSC characteristics" are not supported by the data presented here. Again, I am not convinced that hematopoietic cells can be efficiently generated in this system, and certainly not HSCs or pre-HSCs.

We have provided evidence in the original manuscript, and now through additional experiments, that there is haematopoietic specification, including of progenitor cells, in the haemogenic gastruloid system. Molecular markers are shown in revised Fig. 2F and Fig. 3 and supplements; CFC assays are shown in revised Fig. 2D-E; cytospins are in revised Fig. 2 S2A; further analysis of 4-week implants and new analysis of 8-week implants (discussed below) are in revised Fig. 4 B-D and Fig. 4 S2 and we discussed the new scRNA-seq projections above. Importantly, we have never claimed, and again do not, that haemogenic gastruloids generate HSC. We accept the Reviewer’s comment that we have not provided sufficient evidence for the specification of pre-HSC-like cells and accordingly now refer more generically and conservatively to progenitors.

FACS analysis in 3A is again very unconvincing. I do not think the population identified as C-Kit+ CD144+ is real. Also, why not try gating the other way around, as commonly done (e.g. VE-Cad+ Kit+ and then CD41/CD45)?

Our gating strategy is not unconventional, which was done from a more populated gate onto the less abundant one to ensure that the results are numerically more robust. In the case of haemogenic gastruloids, unlike the AGM preparations the Reviewer may be referring to, CD41 and CD45+ cells are more abundant as there is no circulation of more differentiated haematopoietic cells away from the endothelial structures. This said, we did perform the gating as suggested (Rev Fig. 2), indeed confirming that most VE-cad+ Kit+ cells are CD45+. Interestingly VE-cad+Kit- are predominantly CD41+, reinforcing the haematopoietic nature of these cells.

The authors must have tried really hard, but the lack of short- or long-engraftment in a number of immunodeficient mouse models (lines 305-313) really suggests that no blood progenitors are generated in their system. I am not familiar with the adrenal gland transplant system, but it seems like a very non-physiological system for trying to assess the maturation of putative pre-HSCs. The data supporting the engraftment of these mice, essentially seen only by PCR and in some cases with a very low threshold for detection, are very weak, and again unconvincing. It is stated that "BFP engraftment of the Spl and BM by flow cytometry was very low level albeit consistently above control (Fig. S4E)" (lines 337-338). I do not think that two dots in a dot plot can be presented as evidence of engraftment.

We have presented the data with full disclosure and do not deny that the engraftment achieved is low-level and short-term, indicating incomplete maturation of definitive haematopoietic progenitors in the current haemogenic gastruloid system. Indeed, by not wanting to overstate the finding, we were deliberately conservative in our representative flow cytometry plots and focused on the PCR for sensitivity. We now present the full flow cytometry analysis for spleen where we preserved more cells after the genomic DNA extraction (revised Fig. 4C) and call the Reviewer’s attention to the fact that detection of BFP+ cells by PCR and flow cytometry in the recipient animals is consistent between the 2 methods (revised Fig. 4C and D; full gels previously presented now in Fig. 4S2C; sensitivity analysis was also previously available and is now in Fig. 4S2B). In addition, we have now also been able to detect low-level myelo-lymphoid engraftment in the bone marrow and spleen 8 weeks after adrenal implantation, again suggesting the presence of a small number of definitive haematopoietic progenitors that potentially mature from the 3 haemogenic gastruloids implanted (Fig. 4E and 4 S2F-G in the revised manuscript. We rephrased Results and Discussion at lines 359-414 and 589-621, respectively, to rectify the nature of the engraftment which we attribute to progenitors.

(7) Given the above, I find that the foundations needed for extracting meaningful data from the system when perturbed are very shaky at best. Nevertheless, the authors proceed to overexpress MNX1 by LV transduction, a system previously shown to transform fetal liver cells, mimicking the effect of the t(7;12) AML-associated translocation. Comments on this section:

The increase in the size of the organoid when MNX1 is expressed is a very unspecific finding and not necessarily an indication of any hematopoietic effect of MNX1 OE.

We agree with the Reviewer on this point; it is nevertheless a reproducible observation which we thought relevant to describe for completeness and data reproducibility.

The mild increase of cKit+ cells (Figure 4E) at the 144hr timepoint and the lack of any changes in CD41+ or CD45+ cells suggests that the increase in Kit+ cells % is not due to any hematopoietic effect of MNX1 OE. No hematopoietic GO categories are seen in RNA seq analysis, which supports this interpretation. Could it be that just endothelial cells are being generated?

The Reviewer is correct that the MNX1-overexpressing cells have a strong endothelial signature, which is present in patients (revised Fig. 5A). We investigated a potential link with C-Kit by staining cells from the replating colonies during the process of in vitro transformation with CD31. We observed that 40-50% of C-Kit+ cells (20-30% total colony cells) co-expressed CD31, at least at early plating. These cells co-exist with haematopoietic cells, namely Ter119+ cells, as expected from the YSlike erythroid and EMP-like affiliation of haematopoietic output from 144h-haemogenic gastruloids. These data are included in Fig. 6S1A-B (in text 506-507) of the revised manuscript.

(8) There seems to be a relatively convincing increase in replating potential upon MNX1-OE, but this experiment has been poorly characterized. What type of colonies are generated? What exactly is the "proportion of colony forming cells" in Figures 5B-D? The colony increase is accompanied by an increase in Kit+ cells; however, the flow cytometry analysis has not been quantified.

Given the inability to replate control EV cells, there is not a population to compare with in terms of quantification. The level of C-Kit+ represented in Fig. 6E of the revised manuscript is achieved at plate 2 or 3 (depending on the experiment), both of which are significantly enriched for colony-forming cells relative to control (revised Fig. 6B, D).

(9) Do hGx cells engraft upon MNX1-OE? This experiment, which appears not to have been performed, is essential to conclude that leukemic transformation has occurred.

For the purpose of this study, we are satisfied with confirmation of in vitro transformation potential of MNX1 haemogenic gastruloids, which can be used for screening purposes. Although interesting, in vivo leukaemia engraftment from haemogenic gastruloids is beyond the scope of this study.

Reviewer #2 (Recommendations for the authors):

(1) Minor comments

(a) I find the denomination "hGx" very confusing as it would suggest that these gastruloids are human, whereas, in fact, they are murine.

We agree with the Reviewer on the confusing nomenclature and have edited the manuscript to call “haemGx” instead.

(b) I find the presence of mast cells in CFC of MNX1-OE cultures very puzzling as this does not bear any resemblance to human leukemia.

We detect an enrichment of mast cell transcriptional programmes, as defined by the cell type repositories. While it is not mast cells to represent leukaemic cells in patients, this ontology is likely to reflect the developmental stage and origin of progenitors which are affected by MNX1.

(2) I have a few suggestions to improve figures and tables clarity, to help readers better follow the data presented.

(a) To enhance readability, it would be beneficial to highlight the genes mentioned in the text within the scRNA-seq figures. Many figures currently display over 30-40 genes in small font sizes, making it difficult to quickly locate specific genes discussed in the text. Additionally, implementing a colorcoding system to categorize these genes according to their proposed lineages would improve clarity and organization.

We have now performed major re-organisation and re-analyses of the scRNA-seq data, which we believe has improved the readability and clarity of the corresponding sections of the manuscript.

(b) The data presented in Supplementary Table 1, along with other supplementary tables, are challenging to interpret due to insufficient annotations. Enhancing these tables with clearer and more detailed annotations would significantly improve clarity and aid readers in understanding the supplementary materials.

Descriptive text has been added to accompany each Supplementary File to aid in understanding the results reported therein.

Reviewer #3 (Recommendations for the authors):

In addition to what was written in the public review, I would suggest the authors simplify and shorten the text. Currently, a lot of unnecessary detail is included which makes the story very hard to follow. Moreover, the authors should modify the figures to make them more comprehensible, especially for RNA-seq data.

We have significantly re-arranged and shortened parts of the manuscript, particularly by focusing the Discussion. Results presentation has also been improved through additional analysis and graphic representation of the scRNA-seq data, which we believe has improved the readability and clarity.s

Version 2

Abstract

Abstract

Current in vitro models of developmental blood formation lack spatio-temporal accuracy and weakly replicate successive waves of hematopoiesis. Herein, we describe a mouse embryonic stem cell (SC)-derived 3D hemogenic gastruloid (haemGx) that captures multi-wave blood formation, progenitor specification from hemogenic endothelium (HE), and generates hematopoietic progenitors capable of short-term engraftment of immunodeficient mice upon maturation in an in vivo niche. We took advantage of the haemGx model to interrogate the origins of infant acute myeloid leukemia (infAML). We focused on MNX1-driven leukemia, representing the commonest genetic abnormality unique to the infant group. Enforced MNX1 expression in haemGx promotes the expansion and in vitro transformation of yolk sac-like erythroid-myeloid progenitors at the HE-to-hematopoietic transition to faithfully recapitulate patient transcriptional signatures. By combining phenotypic, functional and transcriptional profiling, including at the single-cell level, we establish the haemGx as a useful new model for the study of normal and leukemic embryonic hematopoiesis.

Introduction

Blood formation in the embryo is a multi-stage process that develops across multiple cellular niches, reflecting a complex hierarchy of tissue interactions. It involves sequential production of distinct cell types in so-called hematopoietic waves, separated in time and space (Lacaud and Kouskoff, 2017; Costa, Kouskoff and Lacaud, 2012; Medvinsky, Rybtsov and Taoudi, 2011; Dzierzak and Bigas, 2018). The first wave of hematopoiesis produces unipotent red blood cell and macrophage precursors in the yolk sac (YS) from an angioblast precursor which can also form endothelium; it occurs at mouse embryo day (E)7-E7.5 (Fujimoto et al., 2001; McGrath and Palis, 2005; Moore and Metcalf, 1970). The subsequent wave, known as definitive, produces: first, oligopotent erythro-myeloid progenitors (EMPs) in the YS (E8-E8.5); and later myelo-lymphoid progenitors (MLPs – E9.5-E10), multipotent progenitors (MPPs – E10-E11.5), and hematopoietic stem cells (HSCs – E10.5-E11.5), in the aorta-gonad-mesonephros (AGM) region of the embryo proper (Ivanovs et al., 2011; McGrath et al., 2015; Palis et al., 1999; Medvinsky and Dzierzak, 1996; Medvinsky, Rybtsov and Taoudi, 2011). Definitive blood waves are characteristically specified from a specialised hemogenic endothelium (HE) through a process of endothelial-to-hematopoietic transition (EHT) (Zovein et al., 2008; Li, W. et al., 2005). HSC specification from HE is the last event in hematopoietic development and occurs in characteristic cell clusters found on the ventral wall of the dorsal aorta (Ivanovs et al., 2011; Boisset et al., 2010). All blood cell types enter circulation from their original locations and migrate to the fetal liver (FL) (Ciriza et al., 2013; Ghiaur et al., 2008). Most embryonic blood progenitors are transient and sustain blood production exclusively during embryonic development, through differentiation. In contrast, HSCs make little or no contribution to embryonic and fetal blood: instead, they proliferate (i.e. self-renew) to expand their numbers. At least a subset of HSC migrates to the bone marrow (BM) niche at the end of gestation and balances self-renewal and differentiation to maintain blood production throughout the entire post-natal life (Bowie et al., 2006; Kikuchi and Kondo, 2006; Ganuza et al., 2022; Yokomizo et al., 2022).

Blood progenitors and HSCs can be targeted by genetic alterations that confer clonal advantage and/or drive initiation and development of leukemia. Leukemia is the most common malignancy and the most common cause of cancer-related death in the paediatric age group (Steliarova-Foucher et al., 2017; Dong et al., 2020). Several paediatric leukemias originate in utero, and a subset of these are driven by mutations exclusive to the paediatric group (Masetti et al., 2015; Cazzola et al., 2021). Such paediatric-exclusive mutations target blood cell types restricted to development or depend on signals that are specific to embryonic hematopoietic niches, effectively defining distinct biological entities to adult malignancies (Cazzola et al., 2021; Bolouri et al., 2018; Chaudhury et al., 2018). Dissection of these forms of developmental leukemia is confounded by the inability to robustly identify and isolate their embryonic cells of origin, and/or to fully recapitulate the disease in post-natal cells, which leads to failure in generating relevant targeted therapies (Alexander and Mullighan, 2021). One such example is t(7;12)(q36;p13) acute myeloid leukemia (AML) driven by rearrangement and overexpression of the MNX1 gene locus: MNX1-rearranged (MNX1-r) AML is restricted to infants under the age of 2 and carries a high risk of relapse and poor survival (Ragusa et al., 2023). MNX1 is a developmental gene required for specification of motor neurons and pancreatic acinar cells (Thaler et al., 1999; Harrison et al., 1999). MNX1 is not known to participate in embryonic hematopoiesis, however it has been discordantly described to be expressed in some subsets of adult hematopoietic cells (Deguchi and Kehrl, 1991; Nagel et al., 2005; Wildenhain et al., 2012; Taketani et al., 2008). It is ectopically expressed in AML cells carrying one of a variable set of t(7;12) translocations which place the normally silent MNX1 gene under the control of ETV6 regulatory elements (Weichenhan et al., 2023; Bousquets-Muñoz et al., 2024; Ragusa et al., 2023; Ballabio et al., 2009). ETV6 is required for hematopoietic cell specification from HE via VEGF signalling (Ciau-Uitz et al., 2010), potentially positioning MNX1 within a critical embryonic hemogenic regulatory network.

Faithful in vitro recapitulation of developmental blood cell types and their embryonic niches has the potential to dissect the cellular and molecular mechanisms presiding at initiation of developmental leukemia, reveal prognostic biomarkers, and indicate therapeutic vulnerabilities which are otherwise challenging to test in the embryo. Despite widespread use of embryonic stem (ES) cell and induced pluripotent stem (iPS) cell-based models of blood specification, a system which recapitulates the spatial and temporal complexity of embryonic blood formation within the respective time-dependent niches is still lacking (Dijkhuis et al., 2023). In the last few years, gastruloid models have emerged as robust avatars of early development, self-organising processes of symmetry breaking, elongation, multi-axis formation, somitogenesis and early organogenesis, with remarkable similarity to embryo processes in space and time (van den Brink, Susanne C and van Oudenaarden, 2021; Beccari et al., 2018; van den Brink, Susanne C et al., 2020). Gastruloids are grown individually in a multi-well format that facilitates tracking and screening of developmental processes. Relevant to blood formation, a variant gastruloid culture system, which successful organises a heart primordium (Rossi et al., 2021), was also able to recapitulate YS blood formation with erythroid and myeloid cellularity (Rossi et al., 2021; Rossi et al., 2022).

Herein, we describe a new protocol of gastruloid formation, i.e. hemogenic gastruloids (haemGx), that captures putative YS-like and AGM-like waves of definitive blood progenitor specification from HE. The earlier wave of putative YS-like hematopoiesis includes EMP-like cells which further mature in the gastruloid culture. HaemGx AGM-like hematopoiesis captures the emergence of intra-aortic cluster-like cells, including cells capable of short-term multi-lineage hematopoietic engraftment of the spleen and BM of immunodeficient mice. We harnessed the embryonic hematopoietic context provided by haemGx to attest its utility in modelling leukemias with developmental origins, specifically by recapitulating MNX1-r infant leukemia by proxy of MNX1 overexpression. MNX1-overexpressing haemGx showed expansion of cells at the HE-to-EMP transition, which became susceptible to transformation in vitro and recapitulated MNX1-r AML patient signatures, placing MNX1’s leukemogenic effects at a very early stage of developmental blood formation.

Results

HaemGxs capture time-dependent emergence of endothelium and hematopoietic cells

With the aim of recapitulating embryonic hematopoiesis with spatial and temporal accuracy, we modified existing gastruloid formation protocols to (1) promote the specification of hemogenic mesoderm derivatives, and (2) extend developmental time beyond the E8.0-equivalent of early developmental gastruloids (Van den Brink, Susanne C et al., 2014) and the E9.0 timepoint of cardiogenic gastruloids (Rossi et al., 2022). We established a 216h haemGx protocol (Fig. 1A) using a Flk1-GFP mouse ES cell line (Jakobsson et al., 2010), which allows the tracking of early hemato-endothelial specification by expression of the Flk1 (i.e. Kdr) locus. The haemGx protocol includes an indispensable pulse of Activin A to induce hemato-endothelial programmes via TGF-β signalling (visible by proxy of GFP-tagged Flk1 activation), in addition of the characteristic gastruloid patterning via WNT activation by CHI99021 supplementation at 48h (Fig. 1S1A-B). We promoted hemato-endothelial programs through addition of VEGF and FGF2 at 72h (Sroczynska et al., 2009) (Fig. 1A). We observed extension of a polarised Flk1-GFP-expressing branched endothelial network (Fig. 1B). Flow cytometry analysis revealed the onset of VE-cadherin+ C-Kit+ cells suggestive of EHT, with a peak at 120-144h (Fig. 1C-D). This was followed by a surge in CD41+ candidate hematopoietic cells at 144h (Fig. 1C-D). In our experience, observation of early polarised activation of the Flk1 locus at 96h (Fig. 1B) was predictive of CD41+ detection 2 days later, suggesting that TGF-β activation in early patterning is required for later hematopoietic development, putatively by modifying mesoderm specification. We incorporated a 24-hour pulse of sonic hedgehog (Shh) between 144-168h to mimic the aortic patterning that precedes cluster formation (Fig. 1A) (Rybtsov et al., 2014). Continued exposure to VEGF after the Shh pulse resulted in specification of CD45+ hematopoietic cells progenitors (Fig. 1C-E), which were apparent after 168h. The proportion of CD45+ cells was enhanced by the addition of SCF, FLT3L and TPO (Fig. 1S1C-D), a combination of cytokines routinely used for HSC maintenance; conversely, it did not require continued addition of FGF2 after 168h (Fig. 1S1E), which was omitted. CD45+ cells, some of which co-expressed C-Kit, were present in a small number of ‘cluster’-like structures adjacent to Flk1-GFP+ Kit+ endothelium (Fig. 1E). The process is overall reminiscent of hematopoietic progenitor cell budding from the aortic HE, suggesting recapitulation of hemogenic tissue architecture. Time-dependent specification of C-Kit+, CD41+ and CD45+ cells in haemGx could be obtained with other mouse ES cell lines, including commonly used E14Tg2a (E14) cells (Fig. 1S1F-H), with successful specification of comparable frequencies of CD45+ progenitors (Fig. 1S1I), attesting to the reproducibility of the protocol across cell lines of different genetic backgrounds.

Hemogenic gastruloids (haemGx) produced from mES cells promote hemato-endothelial specification with spatio-temporally accurate ontogeny.

(A) Timeline of mES cells assembly and culture into gastruloids over a 216h period with the addition of specified factors for the promotion of hemato-endothelial specification. The 24-hour pre-treatment in 2i+LIF was omitted when mES cultures showed no signs of differentiation. (B) Imaging of haemGx over time from 72h to 216h at 10x magnification, showing the assembly and growth of the 3D structures and the polarization of the Flk-1-GFP marker from 96h; scale bar: 100mm. (C) Flow cytometry timecourse analysis of haemGx for the presence of CD144 (VE-cadherin), C-Kit, CD41, and CD45 markers; representative plots. (D) Quantification of analysis in (C). Violin plots with median and interquartile range of up to 26 replicates; Kruskal-Wallis CD144 (p=0.0755), C-Kit (p<0.0001), CD41 (p<0.0342), CD45 (p<0.0001). Shown are significant Dunn’s multiple comparisons tests; adjusted q-value. (E) Immunostaining of whole individual haemGx at 216h showing the localized expression of Flk1-GFP (green), C-Kit (yellow), and CD45 (red) markers; nuclear staining is in DAPI (blue); scale bar: 100μm.

Overall, our extended haemGx model could capture time-dependent specification of CD41+ and CD45+ hematopoietic cells, with recapitulation of the cluster-like spatial arrangement of CD45+ cell emergence from HE, thus showing initial promise as a tool for the exploration of normal and perturbed developmental hematopoiesis.

HaemGx specify cells with phenotypic and functional hematopoietic progenitor characteristics

We confirmed the induction of endothelial patterning from the Flk1-GFP population by immunofluorescence staining showing the branched vascular-like network of Flk1 co-expressing CD31 (Fig. 2A). Endothelial identity of Flk1-GFP+ populations was also detectable by flow cytometry through the presence of endothelial markers CD31 and CD34 (Fig. 2S1A-B). Given the co-expression of CD31 and CD34 with hematopoietic progenitors emerging from HE, we inspected CD41+ and CD45+ cells for the presence of these markers. Indeed, we found that subsets of CD41+ and CD45+ hematopoietic cells co-express CD31 and CD34 (Fig. 2B, Fig. 2S1C-D), with increased relative representation of hemato-endothelial markers within CD41+ cells at 144h and CD45+ at 216h (Fig. 2C), consistent with two successive waves of hemogenic emergence from HE. CD45+ cells from 216h haemGx also co-expressed C-Kit, Flk1-GFP and VE-cadherin, a phenotype enriched in emergent hematopoietic stem and progenitor cells (HSPC) in the AGM (Fig. 2S1E-F). The phenotype is detected in 1.8% of CD45+ cells on average (Fig. 2S1G), corresponding to an estimated 8 cells per haemGx. In agreement, analysis of the haemGx progenitor content at 216h using multipotential CFC assays (Fig. 2D-E) revealed the presence of >50 CFC/haemGx, of which an average of 7 have mixed-lineage granulocytic-monocytic-erythroid (GEM) potential, numerically matching the estimated number of HSPC. Other progenitors were predominantly granulo-monocytic (GM) and erythroid (E) progenitors (Fig. 2D-E). Analysis of the cellularity of dissociated haemGx at the 216h timepoint using Giemsa-Wright stained cytospins revealed differentiated cells of G, M, E and megakaryocytic lineages (Fig. 2S2A), putatively reflecting the progeny of EMP-like cells.

HaemGx produce hematopoietic output following the emergence of specialised endothelium.

(A) Immunostaining of whole individual haemGx at 144h showing the establishment of a vascular network by co-expression of CD31 and Flk1-GFP; nuclear staining is in DAPI (blue); scale bar: 100μm. (B) Flow cytometry analysis of haemGx cells at 144h and 216h stained for CD41, CD45, CD31, and CD34. (C) Quantification of flow cytometry analysis in (B) of CD41+ and CD45+ fractions co-expressing CD34 and CD31. Mixed effect analysis with a Ŝidàk test for multiple comparisons. (D) Colony-forming cell (CFC) assay of 216h-haemGx in multipotential methylcellulose-based medium. GEM: granulocyte-erythroid-monocyte; GM: granulocyte-monocyte; M: monocyte; E: erythroid. Please note that the medium does not include TPO and does not assess the presence of megakaryocytic progenitors. CFC frequency in 3 haemGx, n=3, mean±SD. (E) Representative photographs of colonies quantified in C; scale bar=2mm. (F) Real-time quantitative (qPCR) analysis of expression of hemato-endothelial genes by timepoint. Relative gene expression fold change calculated by normalization to Ppia. Bars represent mean of 3 replicates and show individual data points; p values by ANOVA across datapoints; post-hoc statistical significance between specific variables by Tukey’s test and shown by brackets.

We observed temporal progression of expression of key markers by real-time quantitative PCR (qPCR), configuring a sustained hemato-endothelial program from 120h through 216h (Dll4 and Mecom), and the generation of hematopoietic populations based on the time-dependent patterns of Myb, Runx1, Hlf and Hoxa9 (Fig. 2F). We note that Myb peaks at 144h, coincident with the emergence of CD41+ cells, which we postulate to have YS-like EMP characteristics. On the other hand, Runx1 has a bimodal pattern peaking at 144h, and later at 216h, suggesting the presence of two distinct hematopoietic waves. The later timepoint of 216h has significantly higher expression of both Hoxa9 and Hlf raising the possibility that the second wave indeed corresponds to the specification of AGM-like HSPC. In order to further test the capture of YS-like and AGM-like hematopoietic waves, given the lack of markers that specifically associate with AGM hematopoiesis, we took advantage of the distinct regulatory requirements of polycomb-associated EZH2, which is necessary for EMP differentiation in the YS, but not for AGM-derived hematopoiesis or primitive erythroid progenitors (Neo et al., 2018; Neo et al., 2021). Upon EZH2 inhibition via treatment with GSK216 from 120h onwards, we observed a reduction of CD41+ cells selectively at 144h, with no change in CD45+ cells at the later timepoint of 216h (Fig. 2S2B-C). This suggests that the cells present at the later timepoint may not be the progeny of the former, thus supporting recapitulation of 2 independent hematopoietic waves, the latter putatively AGM. Additionally, the vulnerability of 144h CD41+ to EZH2 inhibition suggests that they include YS-like EMP and do not predominantly reflect EZH2-independent primitive YS hematopoiesis.

HaemGx progenitors capture time-dependent signatures of YS-like and AGM-like hematopoiesis

To better characterize the extent and progression of developmental hematopoiesis in haemGx, we performed single-cell RNA-sequencing (scRNA-seq) time-course analysis of haemGx cell specification. We sorted cells from 2 independent haemGx cultures at 120, 144, 168, 192 and 216h, and profiled a total of 846 cells using the Smart-Seq2 protocol (Picelli et al., 2014) (Fig. 3S1A). We analyzed unfractionated live single haemGx cells at 120-192h and enriched the dataset with sorted CD41+ cells at their peak time of 144h, and sorted CD45+ cells at 192 and 216h. At timepoints of significant enrichment in CD41+ (144h) and CD45+ cells (192h) (Fig. 1D), we also sorted C-Kit+ cells. Library preparation and sequencing generated an average of 120000 reads/cell, which were mapped to an average of 4000 genes/cell, with minimal signs of cell stress / death as measured by mitochondrial DNA fraction (Fig. 3S1B). Read and gene counts were similar between biological replicates, cell types and at different timepoints (Fig. 3S1C). The exception was 120h unfractionated haemGx cells, which despite similar sequencing depth (average read counts), were mapped to twice the number of genes (Fig. 3S1C), possibly reflecting multi-gene program priming at the onset of hemogenic specification. We selected highly varying genes (HVG) before principal component analysis (PCA) dimensionality reduction and retained the most relevant dimensions. In the PCA reduced space, we constructed a KNN graph and used uniform manifold approximation and projection (UMAP) to visualize the data on 2 dimensions, looking for cell communities using Leiden clustering (Fig. 3S1D). We identified 12 cell clusters of which 2 (clusters 0 and 5) almost exactly mapped to C-Kit+ cells (also positive for Sca-1), 2 (clusters 1 and 8) contained CD45+ sorted cells, and cluster 4 uniquely captured the CD41+ cells observed at 144h (Fig. 3S1D). Although some unfractionated cells overlapped with the hemogenic cell clusters (Fig. 3S1D, middle panel), most occupied different transcriptional spaces, reflecting the relative frequency of hematopoietic cells and suggesting the presence of potential hemogenic niches.

Time-resolved analysis of haemGx by scRNA-seq captures successive waves of hematopoietic specification.

(A) UMAP clustering of subsets of sequenced cells expressing CD41, C-Kit, and CD45 at 144h, 192h and 216h, annotated by timepoint (left panel), sorting condition (middle panel), and Leiden clustering (right panel). (B) UMAP panels highlighting the expression of specific hemato-endothelial markers in the transcriptional spaces defined in (A). Color scale indicates expression levels in counts. (C) Heatmaps of differentiation of an arterial endothelial programme in haemGx from scRNA-seq data of sorted on expression of C-Kit at 144h and 192h. Color scale indicates expression in counts. (D) Violin plots quantifying the expression of hemato-endothelial markers in C-Kit+ fraction from clusters in (A) comparing 144h and 192h timepoints. Wilcoxon test; significant p value<0.05. (E) Expression of definitive hematopoietic genes in in C-Kit+, CD41+ and CD45+ cells at 144h and 192/216h timepoints; expression levels as normalised counts per million reads. Mann-Whitney or Kruskal-Wallis with Dunn’s multiple comparisons; significant p or q-value<0.05.

To explore tissue and lineage affiliation of the different clusters, we performed differential gene expression against all other cells using Wilcoxon ranking test, and established cluster classifier gene lists (Supplemental File S1). Top classifier genes for clusters 0 and 5 had endothelial affiliation (Supplemental File S1-S2), and the cellular composition of the clusters broadly reflected their time signature (Fig. 3S1D): cluster 5 comprised 144h cells and cluster 0 included most C-Kit+ cells at 192h. Cluster 5 occupied the vicinity of 144h CD41+ cells (Fig. 3S1D), which carried an erythroid-biased signature (Supplemental File S2). On the other hand, 192/216h-CD45+ cells in cluster 8 expressed myeloid and lymphoid genes (Supplemental File S1-S2). The remaining clusters largely corresponded to time-restricted unfractionated cells (Fig. 3S1D, bottom panel), with clusters of cells present at later timepoints matching mesenchymal stromal cell and autonomic neuron signatures (Fig. 3S1E), potentially configuring incipient formation of niches relevant to production of HSC (Kapeni et al., 2022; Fitch et al., 2012).

To specifically explore the extent to which haemGx capture putative YS and AGM-like hematopoiesis, we re-clustered cells sorted on hemato-endothelial markers CD41, C-Kit, and CD45 at the 144h, 192h and 216h timepoints (Fig. 3A). We mapped the expression pattern of endothelial (Pecam and Cd34), HE (Mecom and Cd44) and hematopoietic (Runx1, Mllt3, Hoxa9, and Hlf) markers across the time-dependent clusters (Fig. 3B). We observed a coincidence of Pecam, Cd34, and Mecom expression with clusters 0 and 3, which capture C-Kit+ cells at 144 and 192h, respectively. Some positive cells were also seen within CD41+ and CD45+-enriched clusters 2, and 1 and 4, respectively, supporting the presence of phenotypic progenitors, as suggested by flow cytometry analysis (see Fig. 2). CD41+ and CD45+-sorted clusters, on the other hand, show expression of Runx1, confirming their hematopoietic affiliation. Cd44, which is upregulated at the endothelial-to-hematopoietic transition (EHT) and in association with the AGM region (Oatley et al., 2020; Fadlullah et al., 2022), is detected predominantly in C-Kit+-enriched cluster 3 and CD45+-sorted clusters 1 and 4, corresponding to the later timepoints of 192 and 216h. This supports the notion that the later timepoints may capture AGM-like EHT, which is also suggested by significant increases in expression of VE-cadherin (Cdh5), Sox17, as well as Cd34, Cd44 and Procr (Fadlullah et al., 2022; Jung et al., 2019; Oatley et al., 2020) in 192h, endothelial-enriched, C-Kit+ cells (Fig. 3C-D). Important in this context, these cells express arterial endothelial markers, including Flt1, Efnb2 and Dll4 steadily at both timepoints (Fig. 3C and 3S2B). Hoxa9 expression also associates with arterial specification and definitive AGM-like hematopoiesis: albeit at low frequency, it is detected exclusively in C-Kit and CD45-sorted cells in the 192 and 216h-timepoints of haemGx differentiation (Fig. 3B, E). These cells also express Mllt3 and, rarely, Hlf, compatible with incipient specification of an AGM-like definitive blood program (Fig. 3B, E). Mllt3 is additionally expressed in CD41+-sorted cells of cluster 2, in line with its role in early erythroid and megakaryocytic lineage identity (Pina et al., 2008); Myb is also more frequently expressed in this population (Fig. 3E). In contrast, Ikzf1, which regulates lymphoid specification (Huang et al., 2019), gradually increases its expression, and indeed frequency, in 216h-CD45+ sorted cells. A handover from erythroid/megakaryocytic to myelo-lymphoid lineage affiliation is apparent between 144h and 216h haemGx blood cells (Fig. 3S2C). Of note, a small number of erythroid-affiliated cells at 216h express adult globins, suggesting a definitive origin (Fig. 3S2C).

Altogether, scRNA-seq analysis support the suggestion from phenotypic and in vitro functional data, that the haemGx model captures two waves of hematopoietic specification from HE-like cells and may configure pre-definitive and definitive embryonic blood formation.

Late-stage haemGx cells transcriptionally resemble mouse AGM progenitors and have short-term engraftment potential

To investigate the putative YS-like pre-definitve and AGM-like definitive nature of haemGx-generated hematopoietic progenitors, we conducted projections of haemGx sc-RNAseq data onto bona fide embryonic populations. We considered 3 distinct scRNA-seq datasets, which (1) capture arterial and haemogenic specification in the para-splanchnopleura (pSP) and AGM region between E8.0 and E11 (Hou et al. 2020) (Fig. 4S1A); (2) uniquely capture YS, AGM and FL progenitors and the AGM EHT (Zhu et al., 2020) (Fig. 4S1B); and (3) interrogate the mouse AGM at the point of HSC emergence from the dorsal aorta HE (Thambyrajah et al., 2024) (Fig. 4S1C). We extracted highly-varying genes from each of the 3 datasets to construct dimensionality reduction maps of the data (Fig. 4S1A-C), and conducted k-Nearest Neighbor (KNN) analysis to classify and regress our haemGx scRNA-seq data onto those maps considering 3 neighbors (Fig. 4A). HaemGx cells with correlation metric >0.7 were projected onto the respective embryonic datasets (Supplemental File S3). Projections onto Hou et al. (2020) positioned clusters 0 and 3, corresponding to endothelium, including arterial endothelium and HE-like cells, between days E8.5 and E10, putatively spanning YS-like and AGM-like blood formation. A subset of CD45+-sorted cells, as well as a small numbers of CD41+ cells from cluster 2 projected onto E10 HE and intra-aortic cluster (IAC) cells, suggesting the presence of definitive HSPC or their precursors in haemGx (Fig. 4A, Supplemental File S3). We sought to further check the transcriptional affiliation of haemGx with intra-embryonic, definitive hematopoiesis, by considering the dataset by Zhu et al. (2020), which uniquely captures different hematopoietic locations in the same study (Fig. 4A). We matched the identity of a large proportion of CD41+-sorted cells to YS EMP, with some of the same cluster 2 cells mapping to erythroid progenitors, and some later CD45+-sorted cells resembling myeloid progenitors (Fig. 4A, Supplemental File S3). On the other hand, a subset of CD45+-sorted cells mapped to IAC and lymphoid-myeloid primed progenitors (LMPP) in the FL (Fig. 4A, ), compatible with the presence of AGM-like definitive HSPC in haemGx. Interestingly, some CD41+-sorted cells also map to IAC, and indeed to HE, suggesting that this earlier population may contain HSPC precursors, in addition to YS-like progenitors. IAC and HE-like cells were also found within C-Kit+ cells, mostly from the 192h timepoint, while the majority of C-Kit+ cells captured different endothelial identities, as expected from their transcriptional profiles. Projection onto Thambyrajah et al. (2024) (Fig. 4A, Supplemental File S3), conflates YS-like and AGM-like progenitors onto the candidate HSC-enriched embryo AGM cluster 2, highlighting the similarity between transcriptional programs at different sites of embryonic blood emergence. CD45+ cells are more distally-projected onto the same cluster, as shown for E11.5 cells in the Thambyrajah et al. (2024) study, some of which correspond to the candidate FL-LMPP in Zhu et al. (2020) (Supplemental File S3). HaemGx cell similarities to Thambyrajah et al. (2024) HSC-enriched cluster 2 also include HE-like and IAC-like cells (as per Zhu et al. 2020 projections), mostly E9.5-like as per Hou et al. (2020), overall suggesting putative early specification of the AGM region within haemGx.

Late-stage haemGx contain hematopoietic output with transcriptional alignment to mouse embryonic populations and in vivo engraftment potential.

(A) Projection of clustered haemGx cells (CD41, C-Kit, and CD45 at 144h, 192h and 216h; Fig. 3A) onto mouse single-cell RNA-seq datasets capturing: arterial and haemogenic specification in the para-splanchnopleura (pSP) and AGM region between E8.0 and E11 (Hou et al. 2020); YS, AGM and FL progenitors and the AGM EHT (Zhu et al., 2020); HSC emergence from the dorsal aorta HE (Thambyrajah et al., 2024). Gastruloid cell projection identified as per their cluster of origin (Fig. 3A). (B) Schematic representation of the experimental workflow for execution and analysis of haemGx implantation in the adrenal gland of immunodeficient mice. (C) PCR detection of haemGx genomic (g)DNA in the bone marrow (BM) and spleen (Spl) of immunodeficient mice 4 weeks after unilateral adrenal implantation of 3 gastruloids/gland. Analysis of 5 replicates of 100ng gDNA/recipient tissue; control animal was injected unilaterally with PBS in an adrenal gland in parallel with experimental implantation. Reaction positive control used 100ng of gDNA from Rosa26-BFP::Flkj1-GFP mES cells (mES) used to generate haemGxs. NTC: no template control. (D) Flow cytometry plots of engraftment detection by BFP expression in bone marrow (BM) of recipient mice 4 weeks following implantation of 216h haemGx. (E) Flow cytometry plots of lineage affiliation of BFP+ engraftment in bone marrow (BM) of recipient mice 8 weeks following implantation of 216h haemGx.

Late-stage haemGx cells can engraft hematopoietic tissues upon maturation

The presence of putative AGM-like hematopoietic cells within haemGx led us to investigate whether haemGx-derived cells had the ability to engraft the hematopoietic system of immunodeficient animals. In early experiments, we had transplanted dissociated haemGx cells obtained at 192 and 216h into irradiated immunocompetent C57Bl/6 recipients, sub-lethally irradiated immunodeficient NSG mice, or non-conditioned NSGW41 hosts, but failed to detect signs of engraftment of peripheral blood, spleen (Spl) or bone marrow (BM) at 12 days (including absence of spleen colony formation), at 4-6 weeks (short-term engraftment), or beyond 8-10 weeks (long-term engraftment) by either flow cytometry detection of CD45.2 or by genomic DNA (gDNA) PCR analysis of the Flk1-GFP locus.

Failure to engraft was perhaps not surprising given that transcriptional programs position haemGx no later than E9.5-E10, which precedes HSC specification in the AGM. Additionally, specification of HSC-supporting stroma within haemGx, namely mesenchymal cells (Kapeni et al., 2022) and elements of the sympathetic nervous system (Fitch et al., 2012), is only detectable at the 192h timepoint, suggesting that further 3D maturation in vivo may be required. We thus tested implantation of undissociated haemGx in the adrenal gland of Nude mice (Fig. 4B), a topography more commonly used to support tumor development, but that was recently shown to be able to support HSC development, possibly with contributions from the sympathetic adrenal medullary niche (Schyrr et al., 2023). In order to facilitate the analysis of engrafted animals, and given the coincidence of CD45 isoform between Flk1-GFP mouse ES cells and the Nude mice in which the experiments were performed, we engineered constitutive expression of BFP from the Rosa26 locus in the Flk1-GFP mES cell line (Rosa26-BFP::Flk1-GFP) (Fig. 4S2A). We implanted 3 haemGx each unilaterally in the adrenal gland of unconditioned recipient mice (Fig. 4B). Control mice were injected with an equivalent volume of PBS. Animals were sacrificed at 4 or 8 weeks after implantation, with 4-5 experimental animals and 1 control collected at each timepoint. We checked Spl and BM BFP engraftment, by flow cytometry and by PCR analysis of DNA. Sensitivity of PCR analysis was determined by serial dilution of Rosa26-BFP::Flk1-GFP mouse ES cells in the human K562 cell line, with detection of as low as 10 pg of mouse input in a total of 100 ng gDNA (Fig. 4S2B), corresponding to approximately 3 cells (in 300,000). PCR analysis of BFP engraftment of the Spl and BM 4 weeks after adrenal implantation of 3 haemGx / animal detected BFP amplicon in both tissues in 3 of the 5 samples (#11, 12 and 15) (Fig. 4C and 4S2C). We sampled the gDNA of each recipient 5 times and detected the presence of the amplicon in 3-5 of the replicate samples of the positive animals; replicate sampling of the control was consistently negative (Fig. 4C and 4S2C). The same 3 animals showed BFP detection by flow cytometry above control, but no higher than 0.1% (Fig. 4D). PCR engraftment was nevertheless detectable in both erythroid Ter119+ and lympho-myeloid CD45+ fractions of the recipient animals showed PCR engraftment in both lineage fractions (Fig. 4S2D-E). Analysis of engraftment 8 weeks after implantation showed low-level detection (<0.1%) of BFP+ cells with positivity for CD11b+ myeloid and B220+ lymphoid compartments in BM of 1 recipient (Fig. 4D and 4S2F-G); a 4th recipient had to be sacrificed early due to infection.

Overall, implantation in the adrenal niche and/or preservation of the 3D architecture enabled low-level in vivo hematopoietic activity, although the mechanism remains unclear. In some recipients, hematopoietic engraftment was accompanied by in situ growth and multi-tissue differentiation of the implanted haemGx (Fig. 4S2H). However, adrenal enlargement was not observed in all animals with hematopoietic engraftment (e.g. 4 weeks’ #11 had a normally sized adrenal, as did all animals analysed at 8 weeks), making it unlikely that hematopoietic activity results from teratoma formation by residual pluripotent cells. Late-stage haemGx (216h) may thus contain hematopoietic progenitors with some short-term engraftment potential but incomplete functional maturation, which are present at low frequency. Engraftment is erythro-myeloid at 4 weeks and lympho-myeloid at 8 weeks, reflecting different classes of progenitors, putatively of YS-like and AGM-like affiliation.

HaemGx resolve the early hemogenic ontogeny targeted by infant AML gene MNX1

The ability of haemGx to reconstitute putative YS and AGM-like HE and hemopoietic progenitors positions it as an attractive model to probe cellular origins of developmental leukemia. In particular, forms of AML unique to the first 2 years of life – infant (inf)AML – have been shown to transform fetal liver (FL) but not adult hematopoietic cells (Waraky et al., 2024; Mercher et al., 2009; Chen, W. et al., 2011; Ragusa et al., 2023), compatible with targeting of a transient embryonic cell. In the case of t(7;12) AML (herein MNX1-r) (Ragusa et al., 2023), in which the translocation results in transformation-driving ectopic expression of MNX1 (Waraky et al., 2024), analysis of patient transcriptional signatures (Ragusa et al., 2022) for over-representation of cell atlas-defined affiliations, indicates enrichment in cardiomyogenic, endothelial, HSC/progenitor and mast cells (Fig. 5A). These signature enrichments are compatible with an early hemogenic cell derivative, capturable in the haemGx model. Other cell type-enrichments reflect MNX1 functions in pancreatic and neuronal development (Ragusa et al., 2022) (Fig. 5S1A).

HaemGx with MNX1 overexpression have increased cellularity and enhanced hemogenic endothelial potential.

(A) Cell type enrichment analysis in MNX1-r AML transcriptomes, compared to normal paediatric bone marrow (BM) or other paediatric AML from the TARGET database. GSEA used representative cell type gene sets from the 2021 DB database; bubble plot shows NES scores (color gradient) and statistical significance (– log10(FDR), bubble size). (B) Schematic representation of generation of MNX1-overexpressing and empty vector (EV) mES cells by lentiviral transduction, with subsequent assembly into haemGxs. (C) Imaging of haemGxs with MNX1 overexpression and EV controls at 10x magnification, showing appropriate assembly and polarization of the Flk1-GFP marker. scale bar: 300mm. (D) Size of haemGx at each timepoint, determined by the distance between the furthest extreme points in µm. Mean ± SD of 3 replicate experiments; 2-tailed t-test, p< 0.05 (*), 0.001 (**), 0.0001 (***), and 0.00001 (****). (E-G) Flow cytometry timecourse analysis of (E) C-Kit+, (F) CD41+, and (G) CD45+ cell abundance in MNX1 and EV haemGxs (120-216h). Mean ± standard deviation of 3-7 independent experiments; 2-way ANOVA and Sidak’s multiple comparison test significant at p<0.05 for C-Kit+ cells only (construct contribution to variance p=0.0191; 144h comparison *p=0.0190). (H) Volcano plots of differentially expressed genes (DEGs) by bulk RNA-seq between MNX1 and EV haemGxs at 144h and 216h. Intersection of statistically significant DEGs between 144 h and 216 h (shown in Venn diagram on top right) identified unique DEGs for each condition (labelled in red for 144 h and blue for 216 h). (J) Gene Ontology (GO) term enrichment for differentially upregulated genes between MNX1 and EV haemGx at 144h (red) and 216h (blue), computed in EnrichR using the GO Biological Process repository. (I) Transcription factor (TF) binding site enrichment on differentially upregulated genes between MNX1 and EV haemGx at 144h (red) and 216 hr (blue) using the ENCODE and ChEA Consensus TFs from ChIP database on EnrichR.

We overexpressed MNX1 (MNX1-OE) in Flk1-GFP mouse ES cells by lentiviral transduction (Fig. 5B) and analyzed progression of haemGx generation (Fig. 5C). We used human MNX1 cDNA to distinguish from the endogenous gene, but the degree of homology is nevertheless high (84%), supporting functional equivalence. We confirmed that MNX1 overexpression in haemGx is maintained to endpoint (Fig. 5S1B). MNX1-OE haemGx activated polarized Flk1-GFP expression and elongated with similar kinetics to empty-vector (EV) control (Fig. 5C), albeit with formation of larger haemGx (Fig. 5D) denoting increased cellularity (Fig. 5S1C). From 192h onwards, MNX1-OE haemGx had a higher frequency of spontaneously contractile structures (Fig. 5S1D), compatible with the cardiogenic cell association of patients’ transcriptomes (Fig. 5A).

We interrogated the time-dependent hemogenic cell composition of haemGx by flow cytometry, quantifying C-Kit, CD41 and CD45 (Fig. 5E-G, 5S1E-F), through which we had captured endothelial/HE-like cells, YS-EMP-like, and later putative AGM-like progenitors, respectively. We observed a small but significant expansion of the C-Kit+ compartment specifically at 144h (Fig. 5E, 5S1E). All cell markers were relatively unchanged at later timepoints, although absolute cell numbers were higher in MNX1-OE haemGx (Fig. 5S1C).

To better understand the consequences of MNX1-OE on hemogenic development, we performed RNA-sequencing (RNA-seq) of haemGx at 144h and 216h, capturing putative YS and AGM-like hemopoietic waves, respectively (Fig. 5H). We confirmed expression of the human MNX1 transgene in the sequenced reads (Fig. 5S2A); in contrast, we could not detect endogenous mouse Mnx1 in any of the samples (Fig. 5 S2A), confirming that it does not normally play a role in hemogenic development. DEGs between MNX1-OE and EV (Supplemental File S4) configured distinct MNX1-driven programs at 144h and 216h, with >80% genes timepoint-specific (Fig. 5H). Both programs include up-regulation of hematopoietic-associated genes, capturing EHT and progenitor regulators (e.g. Runx1, and Epor and Zfpm1, respectively) at the earlier timepoint, and more differentiated effectors, e.g. Hba-a1, Hbb-bh1 and Il3ra at 216h, compatible with selective targeting and/or expansion of YS EMP-like cells and their later differentiated progeny (Fig. 5H). An overview of transcriptional regulatory programs at both timepoints denotes a clearer hemogenic enrichment at 144h, with over-representation of GATA2, GATA1 and RUNX1 binding targets (Fig. 5I), putatively capturing an expansion of a HE-to-EMP-like transition, also represented in enriched GO categories related to VEGF signalling and epithelial-to-mesenchymal transition (EMT) (Fig. 5J). In contrast, no enrichment of hemogenic transcription factor (TF)-ChIP targets or related gene ontologies (GO) categories was observed in MNX1-OE haemGx at 216h (Fig. 5I-J), suggesting specific targeting of YS-like hemogenic programs. Instead, there was an enrichment in cardiogenic GO at 216h (Fig. 5J), reflecting the increase in contractile foci in MNX1-OE late haemGx.

We attempted to further deconvolve the cellular heterogeneity underlying bulk RNA-seq DEGs by interrogating the cluster signatures obtained from the single-cell analysis of haemGx development (Fig. 5S2B-C; refer to Fig. 3S1D). In support of an association with hemato-endothelial specification, MNX1-OE transcriptomes showed enrichment of haemGx HE cluster 5 and 0 signatures; the YS-EMP-like cluster 4 was also enriched (Fig. 5S2B). MNX1-OE DEGs at both timepoints also enriched for the relatively less differentiated cluster 10, which predominantly captures unfractionated haemGx cells at 120h, supporting an early ontogenic affiliation. Significantly, clusters 5 and 0 HE signatures were also enriched in MNX1-r infAML patient signatures (Ragusa et al., 2022) (Fig. 5S2B-C), putatively aligning MNX1-OE effects with HE and the hemogenic transition. In contrast, MNX1-OE DEGs specific to 216h showed enrichment in myeloid progenitor / MLP-enriched cluster 8 which captures 192/216h haemGx cells, but diverges from MNX1-r infAML signatures (Fig. 5S2B-C), instead approximating the distinct myelo-monocytic and/or mixed-lineage affiliation of KMT2A-rearranged AML, which is also common in older pediatric and adult patients.

Serial replating of MNX1-OE haemGx cells identifies a candidate C-Kit+ leukemia-propagating cell more susceptible to transformation in the YS period

We further investigated the potential of MNX1-OE haemGx to reflect MNX1-r AML biology by performing serial replating of CFC assays, a classical in vitro assay of leukemia transformation. We compared 144h and 216h haemGx to match functional and transcriptome data, and plated unfractionated haemGx, given the transient nature of C-Kit+ enrichment. Both 144h (Fig. 6A-B) and 216h (Fig. 6C-D) MNX1-OE haemGx replated for at least 5 platings, significantly different to control (EV) (Fig. 6B,D). EV-haemGx replating could not be sustained beyond plate 3 in most experiments started from 216h (Fig. 6C); 144h EV-haemGx exhibited some low-level replating until plate 5 (Fig. 6A), but the colonies were formed by few scattered cells clearly distinct from the well-demarcated colonies obtained from MNX1-OE haemGx. We analyzed early-stage (plate 1) and late-stage (plate 5) colonies by flow cytometry to identify the nature of the replating cells. Cells selected through replating were C-Kit+ (Fig. 6E), matching the cell phenotype transiently enriched at 144h. MNX1-OE C-Kit+ cells co-expressed CD31 (Fig. 6S1A-B), underpinning the endothelial signatures of MNX1-r leukemia. Early replating of MNX1-OE haemGx also enriched for erythroid-affiliated Ter119+ cells, particularly when initiated from the 144h timepoint (Fig. 6S1C), but this was not sustained through replating. Myelo-monocytic-affiliated markers, including CD11b or CD45, were not present (Fig. 6E, 6S1C).

MNX1 selects C-Kit+ clonogenic cells and selectively transforms end-stage haemGx.

(A) Representative photographs of serial replating of colony-forming cell assays initiated from EV or MNX1 cells obtained at 144h of the haemGx protocol. (B) Quantification of colony-replating efficiency of EV and MNX1 144h-haemGx (GX) cells. Mean+SD of n>3 replicates; Kruskal-Wallis with Dunn’s multiple comparison testing at significant q<0.05. (C) Representative photographs of serial replating of colony-forming cell assays initiated from EV or MNX1 cells obtained at 216h of the haemGx protocol. (D) Quantification of colony-replating efficiency of EV and MNX1 216h-haemGx cells. Mean+SD of n=5-8 replicates; Kruskal-Wallis with Dunn’s multiple comparison testing at significant q<0.05. (E) Flow cytometry analysis of hemogenic progenitor C-Kit+ cells and myeloid-affiliated CD11b+ cells at first round of plating of 144h and 216h-haemGx initiated from EV and MNX1 cells; representative plots.

We performed RNA-seq analysis of MNX1-OE haemGx CFC cells obtained through re-plating and integrated the data with the 144h and 216h haemGx timepoints to identify genes enriched through the process of transformation. HsMNX1 reads were increased in MNX1-OE CFC samples, supporting selective expansion of transduced cells (Fig. 7A). We performed hierarchical clustering of all DEGs in pairwise comparisons (Fig. 7B), and used Gene Set Enrichment Analysis (GSEA) of MNX1-r patient signatures up-regulated in comparison to other AML in the infant (0-2) age group, to interrogate the different clusters. Cluster k14 had significant enrichment of MNX1-r signatures against all other AML subtypes (Fig. 7C), with considerable overlap between leading-edge (i.e. significantly-enriched) genes (Fig. 7S1A). Importantly, k14 captures the subgroup of DEGs up-regulated upon CFC replating which are also up-regulated in MNX1-OE haemGx at 144h (Fig. 7B), matching the time-dependent enrichment of the putative MNX1-OE propagating population. Interrogation of cell-type-specific signatures within the k14 MNX1-r leading edge genes against the single-cell PanglaoDB atlas captured similarities to hemogenic and non-hemogenic cell types, including interneuron, and other neural signatures (Fig. 7D), which reflect MNX1 biological functions (Harrison et al., 1999; Thaler et al., 1999; Ragusa et al., 2022). The most highly significant signatures capture early developmental hemogenic components, such as ‘endothelial cells’ and ‘erythroid cell precursors’, reflecting the 144h, putative YS affiliation of MNX1-OE haemGx. The ‘mast cells’ affiliation of MNX1-r AML signatures within the haemGx k14 cluster is also likely to reflect a YS EMP-like affiliation (Chia et al., 2023), and is accordingly aligned with MNX1-OE haemGx isolated at 144h (Fig. 7S1B). Giemsa-Wright-stained cytospins of serially-replated CFC assays initiated with 144h MNX1-OE cells showed a shift from blast-like at plate 3 to precursor cell morphology with some differentiated mast cells at plate 5 (Fig. 7E, S1C). We also observed mast cells in 216h-initiated colonies (Fig.7E), but these were differentiated at an earlier plate, suggesting that MNX1-OE-responsive C-Kit+ progenitors experience maturation between 144h and 216h.

HaemGx with MNX1 overexpression recapitulate MNX1-Acute Myeloid Leukemia patient signatures.

(A) Quantification of human MNX1 transcripts in FPKM units from RNA-seq of GX-EV, GX-MNX1, and CFC at 144 hr and 216 hr. Bars represent mean of replicates. (B) Heatmap comparing the expression of all differentially expressed genes (DEGs) between all conditions (GX-MNX1 144 hr vs GX-EV 144 hr; GX-MNX1 216 hr vs GX-EV 216 hr; CFC MNX1 144 hr vs GX MNX1 144 hr; CFC MNX1 216 hr vs GX MNX1 216 hr) as Z score. Hierarchical clustering by Ward D method on Euclidean distances identifies 14 clusters (k). (C) GSEA plots for gene set extracted from k14 from (B) against MNX1-r patients RNA-seq counts vs MLL, core-binding factors (CBF), or other pediatric AML. (D) Cell type analysis of the intersect of leading edge genes LEGs (n=476) from Fig. S5D in MNX1-r patients (k14 LEGs) (B) using the Panglao DB 2021 database. (E) Representative Giemsa-Wright stained dissociated CFC replating (4th plating) MNX1 haemGx cells from 144h and 216h. Solid black arrows, mast cell precursors; dashed black arrows, mast cells.

Taken together, the data positioned the initiating cellular targets of MNX1-r AML at the YS endothelial-to-EMP transition. More broadly, the analysis supports the potential of haemGx to dissect the ontogeny of specific subtypes of hematological cancers initiated in utero.

Discussion

In this study, we developed and characterized a hemogenic gastruloid (haemGx) model from mouse ES cells, which recapitulates YS-like and AGM-like waves of definitive blood progenitor specification, and has incipient ability to engraft adult hematopoietic tissues short-term upon in vivo maturation. By exploring the haemGx model through enforced expression of the MNX1 oncogene, which is a hallmark of infant AML harbouring the t(7;12) translocation, we identified the YS endothelial-to-hematopoietic transition as the transformation target of MNX1 gene rearrangement, shedding light on to its exclusive association with the infant age group.

Our data positions the haemGx model as an advanced platform to probe physiological and pathological development of the hematopoietic system, that benefits from the time-dependent and topographic organization associated with gastruloid models. We constructed the haemGx model by introducing a pulse of Activin A and extending the protocol under hematopoiesis-promoting cytokines, to reach the equivalent of mouse embryonic day 10 (E10) and achieve sequential acquisition of YS-like and AGM-like blood production from HE, benchmarked by surface markers CD41 and CD45, respectively. In agreement with recent observations from Dias et al. (2025), which highlight a critical balance between WNT-driven posterior and Nodal / TGFβ anterior signals in patterning the midbody axis and endo-mesodermal derivatives, we observed that the combination of Activin A and CHIR99021 was critical for robust Flk1/Kdr activation and the efficient initiation of CD41 and CD45-marked hematopoietic specification. This is in contrast with the hematopoietic output of cardiogenic gastruloids reported by Rossi et al. (2022) which did not require Activin A. Although our strict dependence on Activin A to achieve initial activation of Flk1 programs and specification of a vascular network and hematopoietic programs precluded a side-by-side comparison with the Rossi protocol (Rossi et al., 2022), we note their apparent lower efficiency in blood formation, at least at the comparable timepoints of 144 and 168h, and their bias towards erythroid formation, reflecting a putative YS, and potentially primitive, nature. This bias may also be captured in the reported spleen colony formation; however, given the limited characterisation of the colonies in the manuscript, which does not appraise origin, morphology or cellularity, the nature of the spleen engraftment remains speculative. Importantly, standard gastruloid protocols with only CHIR99021 pulse and no additional cytokines (Beccari et al., 2018) can capture a transcriptional erythrogenic wave.

Despite low frequency engraftment of adult hematopoietic tissues by haemGx cells, we demonstrated donor origin of the engrafted cells by flow cytometry and PCR detection of a BFP transgene up to 8 weeks. Engraftment progressed from erythroid-biased at 4 weeks to myelo-lymphoid at 8 weeks, as observed in progenitor engraftment from adult cells (Chun et al., 2024). Critically, successful engraftment required implantation of undissociated haemGx for further maturation in a supportive microenvironment, in this case provided by the adrenal gland (Schyrr et al., 2023). In this study, we did not dissect the relative contribution of the continued maturation in 3D vs the influence of the niche for implantation, as we opted for performing the transplantation as to maximise both. Future experiments will be necessary to understand possible extrinsic niche contributions, including the sympathetic nervous system we sought to include (Kapeni et al., 2022; Fitch et al., 2012), and how they can be incorporated in the haemGx culture by extension and/or further modification of the protocol. Although we observed continued growth and maturation of the haemGx in situ after implantation, this was not the case in all animals with engraftment. On the one hand, this dissociation excludes the likelihood of the engraftment resulting from teratoma formation from the gastruloid itself. On the other hand, it raises the possibility that the inefficiency of the engraftment is the consequence of a low frequency of cells within each haemGx with HSPC properties, which we can seek to resolve by upscaling of the haemGx culture or further expansion of dissociated HSPC (Ng et al., 2024).

Although the system is shy of achieving HSC maturation in situ, it can mature, to produce low-level engrafting multilineage progenitors, and potentially HSCs. This suggests that access to the relevant hematopoietic molecular programs is preserved. Incipient specification of non-hematopoietic tissues shown to support HSC emergence, including PDGFRA+ mesenchymal (Chandrakanthan et al., 2022) and neuro-mesenchymal (Miladinovic et al., 2024) cells, and neural-crest derivatives (Kapeni et al., 2022; Fitch et al., 2012), occurs in sync with development of AGM-like hematopoiesis, to indicate the presence of cues that with further maturation may facilitate HSPC specification at the appropriate developmental time. Interestingly, through extending the haemGx culture, and in addition to AGM-like hematopoiesis, we also captured the potential to form renal tubular structures upon in vivo maturation, suggesting induction of lateral plate and intermediate mesoderm from which the AGM region derives, and putative reconstitution of the intra-embryonic hematopoietic niche.

Establishment of a tractable in vitro model of hematopoietic development which faithfully recapitulates time-dependent emergence of hematopoietic progenitors in their physiological niche would propel translational applications by enabling high-throughput studies of HSC emergence. Most ESC-based models of blood specification do not achieve concomitant recapitulation of the surrounding microenvironment. Protocols typically remove ESC-derived hemogenic cells from an early embryonic-like niche to direct differentiation through genetic manipulation (Sugimura et al., 2017) and/or empirically-determined growth factor addition (Lis et al., 2017; Vo and Daley, 2015; Fowler et al., 2024; Nafria et al., 2020; Sroczynska et al., 2009; Piau et al., 2023), with minimal probing of intermediate steps. When hematopoietic waves have been specifically inspected, emergence of YS and AGM-like stages were conflated in time (Pearson et al., 2015; Murry and Keller, 2008), suggesting a bias towards intrinsic hematopoietic cell potential, at the expense of cell fate modulation by time-coherent extrinsic physiological cues, affecting the nature of cellular output. The present haemGx system, in contrast, successfully deconvolutes YS-like and AGM-like stages of blood production, including capturing differences in the HE generation, suggesting a level of organisation amenable to further development into an integrated hemogenic niche. Indeed, projections to scRNA-seq datasets of in vivo murine hematopoiesis place our endothelial and hematopoietic outputs at coherent temporal windows.

The haemGx protocol could be further developed to include more supportive structures of tissue organisation, as non-hematopoietic gastruloid models have benefitted from mechanical, or chemo-mechanical, support of artificial ECM embedding (Veenvliet et al., 2020; van den Brink, Susanne C et al., 2020; Dijkhuis et al., 2023), to match temporal coherence and tissue polarisation, with detailed tissue architecture illustrated by somite formation. Prior to HSC emergence in the AGM, critical signals such as SCF and Shh, which we have delivered to the cultures, are secreted by different cells on opposite dorsal-ventral locations (Souilhol et al., 2016), indicating the potential relevance of tissue organisation to improve and potentially extend hematopoietic formation. Moreover, HSPC formation in the AGM is facilitated by circulatory shear stress (North et al., 2009; Diaz et al., 2015; Lundin et al., 2020), suggesting that introduction of flow may improve hemato-endothelial organization and boost blood production.

We attested the utility of haemGx’s to interrogate the biology of leukemias with developmental origins, thanks to a nearly full recapitulation of embryonic hemogenic specification, including critical early stages of HE production and EHT to EMP and MLP/MPPs. We modelled MNX1-r AML, a poorly characterized form of leukemia exclusive to the infant age group (Ragusa et al., 2023). Overexpression of MNX1 in the haemGx model propagated cells bearing transcriptional resemblance to patient MNX1-r AML, overall positioning the haemGx as an attractive tractable model in which to study developmental leukemias. We showed that MNX1-OE transiently expanded a C-Kit+ cell with hemato-endothelial characteristics at the YS-like stage, and could selectively sustain the phenotype through serial replating of CFC assays. Replating CFCs could be initiated beyond the YS-like stage, albeit at the relative expense of an early differentiation state and of the similarity to patient transcriptome, putatively defining a window of susceptibility to MNX1-driven transformation.

Importantly, the haemGx model was able to resolve the cellularity and transcriptional features inferred from patient transcriptomes, with the advantage of a higher resolution in dissecting the contribution of cell-of-origin in the appropriate developmental context. A recent paper using MNX1-OE has shown that MNX1 can initiate a transplantable leukemia in immunodeficient mice by targeting FL, but not BM, cells (Waraky et al., 2024). This suggests that ectopic expression of MNX1 transforms an embryonic cell, rather than depends on an embryonic niche. It also suggests that MNX1-OE can cause leukemia as a single genetic hit, although it cannot be excluded that other mutations are required for, and may indeed have contributed to, full transformation in vivo, as observed in a recent report confirming the in utero origin of the leukemia (Bousquets-Muñoz et al., 2024). However, the mouse model does not address the nature of the embryonic cell targeted by MNX1-OE. Like the MNX1-OE haemGx cells selected through serial replating, MNX1-OE mouse leukemia are also C-Kit+ (Waraky et al., 2024). Interestingly, clusters of genes up-regulated in the mouse MNX1-OE leukemia (Fig. 7S1D – k1, 2, 5, 6, and 9) do not robustly separate MNX1-r from other infAML by GSEA analysis (Fig. 7S1E), and have diverged more from the patient data in their cell-type signature analysis than the original MNX1-OE cells in the FL (Fig. 7S1F), putatively selecting a more generic program of leukemia transformation. With the resolution provided by the haemGx model, we propose that the divergence from MNX1-r infAML signatures increases from MNX1-OE haemGx 144h, to 216h to FL, positioning the origin of the AML at the YS stage, likely at the HE-to-EMP transition. This cell may persist in the haemGx culture system to the 216h timepoint, and migrate to the FL in the embryo, eventually extinguishing through differentiation, as suggested by the MNX1-OE haemGx CFC assays. Persistence of the original MNX1-OE target cell will determine the sensitivity to MNX1-OE, and can be addressed in the haemGx system through lineage tracing. Whether additional mutations, namely a reported association between MNX1-r and tri(19) (Espersen et al., 2018; Ragusa et al., 2023), contribute to prolonging the undifferentiated state of the candidate YS target cell with increased probability of transformation, or otherwise expand the targeted cell in a more classical second-hit fashion, can potentially also be screened in the haemGx system, in combination with transplantation experiments.

Over the last decade, the effort to achieve robust hematopoietic production from PSC, mouse and human, has been centred on the generation of HSC, a critical translational need (Chang et al., 2023; Dijkhuis et al., 2023; Fidanza and Forrester, 2021). There are good indications that this is achievable, both with (Vo and Daley, 2015; Dang et al., 2002) and without (Piau et al., 2023) genetic manipulation. However, success remains episodic, mainly because mechanistic understanding of processes underpinning blood generation during development is limiting, exacerbated by limiting access to human embryonic and fetal material. On-going optimization of gastruloid models and the success at integration of organoid cultures in microfluidic systems (Saorin, Caligiuri and Rizzolio, 2023; Quintard et al., 2024) will allow us to further optimize haemGx culture. In conclusion, the haemGx represents a faithful in vitro reference to replace and complement embryo studies, holding promise to understand the biology of healthy and malignant developmental hematopoiesis and advance blood production for clinical use.

Experimental procedures

Materials

Methods

Cell culture

Flk-1-GFP (Jakobsson et al., 2010), Flk-1-GFP::Rosa26-BFP, Sox17-GFP (Niakan et al., 2010), T/Bra::GFP (Fehling et al., 2003), and E14Tg2A (Hooper et al., 1987) mouse embryonic stem cells (mES) lines were cultured in ES+LIF medium in gelatinized flasks (0.1% gelatin) with daily medium change, as previously described (Turner et al., 2017). The ES+LIF medium contained 500 ml Glasgow MEM BHK-21 (Gibco), 50 ml of fetal bovine serum (FBS, Embryonic Stem Cells tested, biosera, Nuaillé, France), 5 ml GlutaMAX supplement (Gibco), 5 ml MEM Non-Essential Amino Acids (Gibco), 5 ml Sodium pyruvate solution (100 mM, Gibco), and 1 ml 2-Mercaptoethanol (50mM, Gibco). Murine LIF (Peprotech) was added at 1000 U/ml. HEK293T cells were grown in Dulbecco’s modified Eagle medium (Gibco) supplemented with 10% FBS (Gibco) and 1% penicillin/streptomycin (Gibco). All cultures were kept at 37°C and 5% CO2.

Hemogenic gastruloids assembly, culture, and dissociation

mES were maintained in ES+LIF medium and transferred to 2i+LIF (containing Chiron and MEK inhibitor PD03) for 24 hours prior to the assembly into gastruloids. 250 untransduced mES cells (or 400 for GX-MNX1 and GX-EV) were seeded in each well of a U-bottom, cell-repellent 96-well plate (Greiner Bio-One, Stonehouse, UK) in 40 μl of N2B27 medium [Takara Bio or home-made as Cold Spring Harbour Protocols ‘N2B27 Medium’ (2017)]. The plate was centrifuged at 750 rpm for 2 minutes to promote deposition and aggregation of the cells and was then incubated at 37°C, 5% CO2 for 48 hours. After 48 hours, 150 μl of N2B27 medium supplemented with 100 ng/ml Activin A (QKine, Cambridge, UK) and 3 μM chiron (Peprotech) was added to each well. At 72 hours, 150 μl of medium were removed, without disrupting the gastruloids in the wells. 100 μl of N2B27 with 5 ng/ml of Vegf and Fgf2 each (Peprotech) were added to each well. From 72 h to 144 h, each day 100 μl of medium were removed and replaced with N2B27 + Vegf + Fgf2. At 144 h, the medium was further supplemented with Shh at 20 ng/ml. From 168 h to 216 h, the medium was N2B27 + 5 ng/ml Vegf, plus 20 ng/ml mTpo, 100 ng/ml mFlt3l, and 100 ng/ml mScf (Peprotech). To dissociate cells from the gastruloid structures, medium was removed and individual gastruloids were collected using a pipette and precipitated at the bottom of a microcentrifuge tube. The remaining medium was aspirated and the bulk of gastruloids was washed in PBS. 50 μl of TrypLE express was added to pelleted gastruloids to be incubated at 37°C for 2 minutes to dissociate cells.

Animals and implantation

Female athymic nude-Foxn1nu (nu/nu) mice (Envigo, Bresso, Italy) were housed under pathogen-free conditions. In accordance with the “3Rs policy”, experiments were reviewed and approved by the Animal Welfare Body (OPBA) of IRCCS Ospedale Policlinico San Martino and by the Italian Ministry of Health (n. 883/2020-PR).

Intact haemGx (3 per mouse in 10 μL of PBS) were inoculated in the adrenal gland of five-week-old mice. Briefly, mice were anesthetized with a mixture of xylazine (ROMPUN, Bayer) and ketamine (LOBOTOR, Acme S.r.l.) and injected with haemGx, after laparotomy, in the capsule of the left adrenal gland, as previously described (Brignole et al., 2023; Pastorino et al., 2003). All mice survived surgery. Control mice (CTR) received vehicle only. Mice body weight and general physical conditions were recorded daily. One month after haemGx inoculation, mice were sacrificed and spleen and bone marrow (BM) from femurs collected. BM cells and spleens, mechanically reduced to cell suspension, were stored at −80°C until use. Adrenal glands were paraffin-embedded for subsequent IHC studies.

Methylcellulose colony forming assays (CFC)

Disassembled gastruloids cells were plated on Mouse Methylcellulose Complete Media (R&D Systems). Cells were first suspended in Iscove’s Modified Dulbecco’s Medium (IMDM, Gibco) with 20% FBS (Gibco) before addition to the methylcellulose medium. Cells were plated in duplicate 35 mm dishes with 2×105 cells/plate. Plates were incubated at 37°C and 5% CO2 for 10 days, when colonies were scored. For serial replating experiments, cells in methylcellulose were collected and washed in phosphate buffer saline (PBS) to achieve single-cell suspensions and replated as described above. D issociated gastruloid cells from CFC were collected as cell suspensions in PBS and centrifuged onto slides (Shandon Cytospin 2 Cytocentrifuge) at 350rpm for 5-7min, with a PBS-only pre-spin at 1000rpm, 2min. Slides were stained in Giemsa-Wright stain for 30 seconds followed by addition of phosphate buffer pH 6.5 for 5 min.

Immunofluorescence staining

Immunostaining of whole gastruloids was performed as described before (Baillie-Johnson et al., 2015). Briefly, gastruloids were fixed in 4% paraformaldehyde (PFA) dissolved in PBS for 4 hours at 4°C on orbital shaking and permeabilised in PBSFT (10% FBS and 0.2% Triton X-100), followed by one hour blocking in PBSFT at 4°C on orbital shaking. Antibody dilutions were made in PBSFT at 1:200 for primary and 1:500 for secondary antibodies. Antibody incubations were performed overnight at 4°C on orbital shaking, and subjected to optical clearing overnight in ScaleS4 clearing solution. Individual gastruloids were then mounted on glass coverslips by pipetting as droplets in ScaleS4 and DAPI nuclear stain.

Histology and H&E staining

Samples were fixed with 10% formalin at room temperature for 48 hours followed by an ethanol series of ethanol and paraffin; ethanol 70% (90 minutes), ethanol 80% (120 minutes), ethanol 95% (180 minutes), 3 times ethanol 100% of 120, 120 and 180 minutes, 50% ethanol/paraffin (3×120 minutes) and 3 times paraffin (60, 90 and 120 minutes) before embedding in histological cassettes. Blocks were sectioned with a HistoCore BIOCUT Microtome (Leica) at 10 mm and slides were stained with Haematoxylin/Eosin.

Imaging

Images of cultured gastruloids and CFC plates were captured using the Cytation 5 Cell Imaging Multi-Mode Reader (Biotek) plate reader using bright field and FITC channels. ImageJ (Schneider et al., 2012) was used for gastruloid size quantification. Confocal microscopy was performed on LSM700 on a Zeiss Axiovert 200 M with Zeiss EC Plan-Neofluar 10x/0.30 M27 and Zeiss LD Plan-Neofluar 20x/0.4 M27 objective lens. Tissue sections were imaged with a THUNDER Imager 3D microscope system (Leica).

Lentiviral vector packaging and transduction

The lentiviral overexpression vector pWPT-LSSmOrange-PQR was used to clone the MNX1 gene cDNA. The viral packaging vectors pCMV and pMD2.G, described in Pina et al. (2008), were used to assemble lentiviral particles using HEK293T cells via transfection using TurboFect Reagent (Invitrogen). Transduction of mES cells was performed overnight by addition of lentivirus to culture medium and washed the following day (Moris et al., 2018). Transduced cells were sorted for positivity to LSSmOrange.

Flow cytometry

Surface cell marker analysis was performed by staining using the antibodies listed in Key Resources Table. Disassembled gastruloids cells were resuspended in PBS, 2% FBS and 0.5 mM EDTA and stained at a dilution of 1:100 for primary antibodies for 20 minutes at 4°C. When indicated, streptavidin was added at a dilution of 1:200. Analysis was performed on ACEA Novocyte (Agilent) or AttuneNxT (Thermo) analyzers, using the respective software packages. Cell sorting was performed using a CS&T calibrated BDFACS Aria III system (488nm 40 mW, 633nm 20 mW, 405nm 30 mW, and 561nm 50 mW), set with the 100μm nozzle at 20PSI and a 4-way purity mask or, in the case of haemGx adrenal implants, a Beckman Coulter CytoFlex SRT (488nm 50 mW, 633nm 100 mW, 405nm 90 mW, and 561nm 30 mW), set with 100μm nozzle at 15PSI, also using 4-way purity. Single-cell deposition in 96-well plates was performed using single-cell sorting mode. Intact cells were gated on FSC-A vs SSC-A plot, followed by doublet exclusion on FSC-A vs FSC-H and SSC-A vs SSC-H, prior to gating on fluorescent parameters for the markers described in the results.

Single cell RNA sequencing of time-resolved hemogenic gastruloids

Gastruloids were collected at different timepoints of the protocol, disassembled and FACS deposited into 96-well plates, either as unsorted (global) or sorted by CD45, CD41 and C-Kit/ScaI markers (sorted). RNA from the cells were extracted from single cells using Smart-seq2 technology at 500Kb and depth of 151 bases. Sequencing reads were quality-checked using FastQC (v0.11.9). We trimmed the samples using trimGalore! (0.6.6) with a cutoff of 30, clipping 15 base pairs and retaining reads of more than 100 bases. Alignment was performed on STAR (2.7.8a) and Mus musculus annotations from GENCODE vM26. Aligned BAM files were annotated using featureCounts (v2.0.1) and count matrices were computed in python by directly accessing the BAM files with gtfparse packages and collapsing lanes.

For quality control, based on the histogram of counts and multimodality distributions, we set a minimum count threshold of 200000 counts, a minimum threshold of 1000 expressed genes, and a maximum threshold of 20% mitochondrial fraction per cell. We performed the same procedure of the second dataset with a more restrictive threshold of 400000 counts and similar expressed genes and mitochondrial thresholds. Cells that did not pass the quality control metrics were omitted from analysis. We normalized the cells to the mean count number per dataset and applied a plus-one-log transformation of the data before proceeding to the downstream analysis.

Dimensionality reduction was performed on feature selection of the gene space using the function scanpy.highly_varying_genes with default parameters. Selection of principal components in principal component analysis (PCA) was performed by heuristic elbow method. Nearest neighbor analysis was constructed by KNN graphs using a correlation metric and 10 nearest neighbors. Data was projected for low dimensional visualization using the UMAP algorithm with default parameters as implemented in scanpy.tl.umap. We used the leiden algorithm as implemented in scanpy.tl.leiden to partition the data into clusters. In order to assess the election of the resolution parameter we used Newman-Girvan modularity as a metric of clustering quality. Differential expressed genes were computed comparing each cluster against the rest using the Wilcoxon test with Benjamini-Hochberg correction.

Annotation and projection to additional datasets

Raw counts matrices were downloaded and processed following the same pipeline as described above for scRNA-seq of gastruloids to generate UMAP and clustering. We implemented the scmap algorithm to compare our scRNA sequencing of gastruloids with the available datasets. We reduced the dimensionality of the space by selecting highly varying genes from the annotated dataset. Then, we constructed a KNN classifier with correlation metric and computed the nearest neighbors of the target data. If neighbors with correlation metrics below 0.7 default standards, the projected cells were not projected onto any cell from the annotated dataset. To visualize the cells over the UMAP plots of the other datasets, we constructed a KNN regressor with three neighbors and a correlation metric. Confusion matrices were used to visualize the overlap between gastruloid clusters and the embryo dataset assigned clusters using scmap.

Bulk RNA sequencing

Total RNA was extracted from disassembled gastruloid cells at 216 hours. Sequencing was performed on NovaSeq PE150 platform, at 20M paired-end reads per sample. Tophat2 with Bowtie2 were used to map paired-end reads to the reference Mus musculus genome build GRCm39. GENCODE Release M30 (Frankish et al., 2019) was used as the reference mouse genome annotation. Aligned reads were filtered by quality using samtools (Li, H. et al., 2009) with a minimum threshold set at 30 (q30). Transcript assembly and quantification was achieved using htseq (Putri et al., 2022). Differential expression between sample and control was performed by collapsing technical replicates for each condition using Deseq2 (Love, Huber and Anders, 2014) in R environment (Deseq2 library v 1.32.0). The differential expression was expressed in the form of log2 fold change and filtered by false discovery rate (FDR) of 0.05.

Gene list enrichment analyses

Gene ontology (GO) analysis was performed in ExpressAnalyst (available at www.expressanalyst.ca) using the PANTHER or GO Biological Process (BP) repository. GO terms and pathways were filtered by false discovery rate (FDR) with a cut-off of ≤ 0.1 for meaningful association. EnrichR (Xie et al., 2021; Kuleshov et al., 2016; Chen, E. Y. et al., 2013) was used for cell type analysis using the Panglao DB (Franzén, Gan and Björkegren, 2019) databases using a FDR threshold of ≤ 0.1 or p value ≤ 0.01, where specified, as well as transcription factor (TF) binding site enrichment using the ENCODE and ChEA Consensus TFs from ChIP database.

Gene Set Enrichment Analysis (GSEA)

Custom gene signatures were used as gene sets for GSEA analysis (Subramanian et al., 2005) on the GSEA software v4.2.3 on RNA sequencing expression values in counts units. GSEA was ran in 10000 permutations on gene set using the weighted Signal2Noise metric. Enrichment metrics are shown as normalized enrichment score (NES) and filtered by FDR ≤ 0.05. Leading edge genes (LEGs) are genes with a “Yes” values for core enrichment. For AML patient analysis, clinical phenotype and expression data (in counts units) were extracted from the GDC TARGET-AML cohorts in the Therapeutically Applicable Research to Generate Effective Treatments project (TARGET, https://ocg.cancer.gov/programs/target), downloaded from the University of California Santa Cruz (UCSC) Xena public repository (last accessed 31st August 2022). Patient samples were selected according to the reported karyotype to include t(7;12), inv(16), MLL, normal karyotype, and t(8;21). GSEA was performed comparing RNA sequencing counts of t(7;12) samples against pooled AML subtypes (inv(16), MLL, normal karyotype and t(8;21)) as “other AML”.

Real-time polymerase chain reaction (qPCR)

Extracted RNA was reverse-transcribed into complementary DNA (cDNA) using High-Capacity RNA-to-cDNA Kit (Applied Biosystems). QPCR was performed using FastGene 2x IC Green Universal qPCR Mix (Nippon Genetics, Duren, Germany) using primers for human MNX1 (forward 5-GTTCAAGCTCAACAAGTACC-3; reverse 5-GGTTCTGGAACCAAATCTTC-3) (Gulino et al., 2021) and Ppia for endogenous control (forward 5-TTACCCATCAAACCATTCCTTCTG-3; reverse 5-AACCCAAAGAACTTCAGTGAGAGC-3) (Moris et al., 2018), or Taqman gene expression assays (Mm00501741_m1 Myb; Mm00439364_m1 Hoxa9; Mm00723157_m1 Hlf; Mm01213404_m1 Runx1; Mm00444619_m1 Dll4; Mm00491303_m1 Mecom; Mm02342430_g1 Ppia) using Taqman Gene Expression Mastermix (Applied Biosystems). Differential gene expression was calculated using the delta delta Ct (ΔΔCt) method.

Polymerase chain reaction (PCR)

Genomic DNA was extracted using Monarch Genomic DNA Purification Kit (New England Biolabs, Hitchin, UK) using manufacturer’s instructions. PCR on genomic DNA was performed using Phire PCR Master Mix (Thermo Scientific) to amplify the BFP locus using forward primer 5’-GCACCGTGGACAACCATCACTT-3’ and reverse primer 5’-CAGTTTGCTAGGGAGGTCGC-3’.

Statistical analysis

Experiments were performed at least in triplicates, unless specified otherwise. Data are plotted to include standard deviation (+/−SD) between replicates. Statistical significance was set at a threshold of p value < 0.05. Statistical analysis was performed in R environment (version 4.1.3) or using GraphPad Prism 8.0 software and is detailed in respective figure legends.